### 2025-03-06 template.tex
white space before \corr

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[runningheads]{llncs}

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{2}

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\newsavebox\pandoc@box
\newcommand*\pandocbounded[1]{% scales image to fit in text height/width
  \sbox\pandoc@box{#1}%
  \Gscale@div\@tempa{\textheight}{\dimexpr\ht\pandoc@box+\dp\pandoc@box\relax}%
  \Gscale@div\@tempb{\linewidth}{\wd\pandoc@box}%
  \ifdim\@tempb\p@<\@tempa\p@\let\@tempa\@tempb\fi% select the smaller of both
  \ifdim\@tempa\p@<\p@\scalebox{\@tempa}{\usebox\pandoc@box}%
  \else\usebox{\pandoc@box}%
  \fi%
}
% Set default figure placement to htbp
\def\fps@figure{htbp}
\makeatother

\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage[misc]{ifsym}
\newcommand{\corr}{(\Letter)}
\makeatletter

\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Fig.}
\else
  \newcommand\figurename{Fig.}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter

\makeatother

\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Recent Advances in Underwater Basket Weaving Under the Extreme Pressure of the Mariana Trench},
  pdfauthor={André Lauren Benjamin; Calvin Cordozar Broadus Jr.; Antwan André Patton},
  pdfkeywords={First Keyword, Second Keyword, Another Keyword.},
  colorlinks=true,
  linkcolor={black},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\hypersetup{nolinks=true}

\title{Recent Advances in Underwater Basket Weaving Under the Extreme
Pressure of the Mariana Trench}

\titlerunning{Underwater Basket Weaving Under Extreme Pressure}

\author{André Lauren Benjamin\inst{1} \and Calvin Cordozar Broadus
Jr.\inst{2,3} \corr \and Antwan André
Patton\inst{1}\orcidID{0000-1111-2222-3333}}

\authorrunning{A.L. Benjamin et al.}

\institute{Fictional Southern University, Savannah GA
31404, USA  \email{\{a.l.benjamin,a.a.patton\}@fsu.fake} \and Fictional
West Coast University, Long Beach CA
90840, USA  \email{ccb@fwcu.fake} \and Secondary European
Affiliation, Tiergartenstr.
17, 69121 Heidelberg, Germany  \email{lncs@springer.com}}

\begin{document}
\maketitle

\begin{abstract}
The abstract should briefly summarize the contents of the paper in
150--250 words.

\keywords{First Keyword \and Second Keyword \and Another Keyword.}

\end{abstract}

\section{Introduction}\label{sec-intro}

\subsection{Main Contributions}\label{main-contributions}

Please note that the first paragraph of a section or subsection is not
indented. The first paragraph that follows a table, figure, equation
etc. does not need an indent, either.

Subsequent paragraphs, however, are indented.

\section{Related Work}\label{related-work}

\subsection{Basket Weaving}\label{basket-weaving}

\subsubsection{Underwater Basket
Weaving}\label{underwater-basket-weaving}

Only two levels of headings should be numbered. Lower level headings
remain unnumbered; they are formatted as run-in headings.

\paragraph{Underwater Basket Weaving Under Difficult
Circumstances}\label{underwater-basket-weaving-under-difficult-circumstances}

The contribution should contain no more than four levels of headings.
Table~\ref{tbl-headings} gives a summary of all heading levels.

\begin{table}[t]

\caption{\label{tbl-headings}Table captions should be placed above the
tables.}

\centering{

\begin{tabular}{lll}
\toprule
Heading level &  Example & Font size and style\\
\midrule
Title (centered) &  {\Large\bfseries Lecture Notes} & 14 point, bold\\
1st-level heading &  {\large\bfseries 1 Introduction} & 12 point, bold\\
2nd-level heading & {\bfseries 2.1 Printing Area} & 10 point, bold\\
3rd-level heading & {\bfseries Run-in Heading in Bold.} Text follows & 10 point, bold\\
4th-level heading & {\itshape Lowest Level Heading.} Text follows & 10 point, italic\\
\bottomrule
\end{tabular}

}

\end{table}%

\section{Recent Advances from the Mariana
Trench}\label{recent-advances-from-the-mariana-trench}

Displayed equations are centered and set on a separate line.

\begin{equation}\phantomsection\label{eq-linear}{
x + y = z
}\end{equation}

Please try to avoid rasterized images for line-art diagrams and schemas.
Whenever possible, use vector graphics instead (see
Fig.~\ref{fig-duck}).

\begin{figure}[t]

\centering{

\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{example-image-duck.pdf}

}

\caption{\label{fig-duck}A figure caption is always placed below the
illustration. Please note that short captions are centered, while long
ones are justified by the macro package automatically.}

\end{figure}%

\begin{theorem}
This is a sample theorem. The run-in heading is set in bold, while
the following text appears in italics. Definitions, lemmas,
propositions, and corollaries are styled the same way.
\end{theorem}

\begin{proof}
Proofs, examples, and remarks have the initial word in italics,
while the following text appears in normal font.
\end{proof}

For citations of references, we prefer the use of square brackets and
consecutive numbers. Citations using labels or the author/year
convention are also acceptable. The following bibliography provides a
sample reference list with entries for journal articles
\cite{ref_article1}, an LNCS chapter \cite{ref_lncs1}, a book
\cite{ref_book1}, proceedings without editors \cite{ref_proc1}, and a
homepage \cite{ref_url1}. Multiple citations are grouped
\cite{ref_article1,ref_lncs1,ref_book1},
\cite{ref_article1,ref_book1,ref_proc1,ref_url1}.

\section{Experiments}\label{experiments}

\subsection{Experimental Setup}\label{experimental-setup}

\subsection{Experimental Results}\label{experimental-results}

\section{Discussion}\label{discussion}

\section{Conclusion}\label{conclusion}

Of course, authors have complete freedom on how they choose to structure
their paper. Section headers from Introduction up to and including
Conclusions are completely up to the discretion of the authors; use
whichever structure you see fit. Title, Abstract, the credits
environment, and References, however, are mandatory.

\begin{credits}
\subsubsection{\ackname} A bold run-in heading in small font size at the end of the paper is
used for general acknowledgments, for example: This study was funded
by X (grant number Y).

\subsubsection{\discintname}
It is now necessary to declare any competing interests or to specifically
state that the authors have no competing interests. Please place the
statement with a bold run-in heading in small font size beneath the
(optional) acknowledgments,
for example: The authors have no competing interests to declare that are
relevant to the content of this article. Or: Author A has received research
grants from Company W. Author B has received a speaker honorarium from
Company X and owns stock in Company Y. Author C is a member of committee Z.
\end{credits}

\bibliographystyle{splncs04}
\bibliography{bibliography}

\end{document}

### 2025-03-06 template.qmd
done here

--- 2025-03-06 template.tex
@@
-% Options for packages loaded elsewhere
-\PassOptionsToPackage{unicode}{hyperref}
-\PassOptionsToPackage{hyphens}{url}
-\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
-%
-\documentclass[runningheads]{llncs}
+---
+title: Recent Advances in Underwater Basket Weaving Under the Extreme Pressure of the Mariana Trench
+title-running: Underwater Basket Weaving Under Extreme Pressure
+format:
+  quarto-ecml-pdf:
+    keep-tex: true
+  quarto-ecml-html: default
+blind-review: false
+author:
+  - name: André Lauren Benjamin
+    affiliations:
+      - name: Fictional Southern University
+        address: Savannah GA 31404
+        country: USA
+        url: \{a.l.benjamin,a.a.patton\}\@fsu.fake
+        id: south
+    url: https://example.org/
+  - name: Calvin Cordozar Broadus Jr.
+    corresponding: true 
+    affiliations:
+      - name: Fictional West Coast University
+        address: Long Beach CA 90840
+        country: USA
+        url: ccb@fwcu.fake
+      - name: Secondary European Affiliation
+        address: Tiergartenstr. 17
+        city: Heidelberg
+        country: Germany
+        postal-code: 69121 
+        url: lncs@springer.com
+  - name: Antwan André Patton
+    orcid: '0000-1111-2222-3333'
+    affiliations:
+      - ref: south 
+author-running: A.L. Benjamin et al.
+abstract: |
+  The abstract should briefly summarize the contents of the paper in 150--250 words.
+keywords: [First Keyword, Second Keyword, Another Keyword.]
+---
 
-\usepackage{amsmath,amssymb}
-\usepackage{iftex}
-\ifPDFTeX
-  \usepackage[T1]{fontenc}
-  \usepackage[utf8]{inputenc}
-  \usepackage{textcomp} % provide euro and other symbols
-\else % if luatex or xetex
-  \usepackage{unicode-math}
-  \defaultfontfeatures{Scale=MatchLowercase}
-  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
-\fi
-\usepackage{lmodern}
-\ifPDFTeX\else  
-    % xetex/luatex font selection
-\fi
-% Use upquote if available, for straight quotes in verbatim environments
-\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
-\IfFileExists{microtype.sty}{% use microtype if available
-  \usepackage[]{microtype}
-  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
-}{}
-\usepackage{xcolor}
-\setlength{\emergencystretch}{3em} % prevent overfull lines
-\setcounter{secnumdepth}{2}
+## Introduction {#sec-intro}
 
-\providecommand{\tightlist}{%
-  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
-\usepackage{calc} % for calculating minipage widths
-% Correct order of tables after \paragraph or \subparagraph
-\usepackage{etoolbox}
-\makeatletter
-\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
-\makeatother
-% Allow footnotes in longtable head/foot
-\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
-\makesavenoteenv{longtable}
-\usepackage{graphicx}
-\makeatletter
-\newsavebox\pandoc@box
-\newcommand*\pandocbounded[1]{% scales image to fit in text height/width
-  \sbox\pandoc@box{#1}%
-  \Gscale@div\@tempa{\textheight}{\dimexpr\ht\pandoc@box+\dp\pandoc@box\relax}%
-  \Gscale@div\@tempb{\linewidth}{\wd\pandoc@box}%
-  \ifdim\@tempb\p@<\@tempa\p@\let\@tempa\@tempb\fi% select the smaller of both
-  \ifdim\@tempa\p@<\p@\scalebox{\@tempa}{\usebox\pandoc@box}%
-  \else\usebox{\pandoc@box}%
-  \fi%
-}
-% Set default figure placement to htbp
-\def\fps@figure{htbp}
-\makeatother
+### Main Contributions
 
-\usepackage[T1]{fontenc}
-\usepackage{graphicx}
-\usepackage{booktabs}
-\usepackage[misc]{ifsym}
-\newcommand{\corr}{(\Letter)}
-\makeatletter
-
-\AtBeginDocument{%
-\ifdefined\contentsname
-  \renewcommand*\contentsname{Table of contents}
-\else
-  \newcommand\contentsname{Table of contents}
-\fi
-\ifdefined\listfigurename
-  \renewcommand*\listfigurename{List of Figures}
-\else
-  \newcommand\listfigurename{List of Figures}
-\fi
-\ifdefined\listtablename
-  \renewcommand*\listtablename{List of Tables}
-\else
-  \newcommand\listtablename{List of Tables}
-\fi
-\ifdefined\figurename
-  \renewcommand*\figurename{Fig.}
-\else
-  \newcommand\figurename{Fig.}
-\fi
-\ifdefined\tablename
-  \renewcommand*\tablename{Table}
-\else
-  \newcommand\tablename{Table}
-\fi
-}
-\@ifpackageloaded{float}{}{\usepackage{float}}
-\floatstyle{ruled}
-\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
-\floatname{codelisting}{Listing}
-\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
-\makeatother
-\makeatletter
-\makeatother
-\makeatletter
-
-\makeatother
-
-\usepackage{bookmark}
-
-\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
-\urlstyle{same} % disable monospaced font for URLs
-\hypersetup{
-  pdftitle={Recent Advances in Underwater Basket Weaving Under the Extreme Pressure of the Mariana Trench},
-  pdfauthor={André Lauren Benjamin; Calvin Cordozar Broadus Jr.; Antwan André Patton},
-  pdfkeywords={First Keyword, Second Keyword, Another Keyword.},
-  colorlinks=true,
-  linkcolor={black},
-  filecolor={Maroon},
-  citecolor={Blue},
-  urlcolor={Blue},
-  pdfcreator={LaTeX via pandoc}}
-
-\hypersetup{nolinks=true}
-
-\title{Recent Advances in Underwater Basket Weaving Under the Extreme
-Pressure of the Mariana Trench}
-
-\titlerunning{Underwater Basket Weaving Under Extreme Pressure}
-
-\author{André Lauren Benjamin\inst{1} \and Calvin Cordozar Broadus
-Jr.\inst{2,3} \corr \and Antwan André
-Patton\inst{1}\orcidID{0000-1111-2222-3333}}
-
-\authorrunning{A.L. Benjamin et al.}
-
-\institute{Fictional Southern University, Savannah GA
-31404, USA  \email{\{a.l.benjamin,a.a.patton\}@fsu.fake} \and Fictional
-West Coast University, Long Beach CA
-90840, USA  \email{ccb@fwcu.fake} \and Secondary European
-Affiliation, Tiergartenstr.
-17, 69121 Heidelberg, Germany  \email{lncs@springer.com}}
-
-\begin{document}
-\maketitle
-
-\begin{abstract}
-The abstract should briefly summarize the contents of the paper in
-150--250 words.
-
-\keywords{First Keyword \and Second Keyword \and Another Keyword.}
-
-\end{abstract}
-
-\section{Introduction}\label{sec-intro}
-
-\subsection{Main Contributions}\label{main-contributions}
-
-Please note that the first paragraph of a section or subsection is not
-indented. The first paragraph that follows a table, figure, equation
-etc. does not need an indent, either.
+Please note that the first paragraph of a section or subsection is not indented. The first paragraph that follows a table, figure, equation etc. does not need an indent, either.
 
@@
 
-\section{Related Work}\label{related-work}
+## Related Work
 
-\subsection{Basket Weaving}\label{basket-weaving}
+### Basket Weaving
 
-\subsubsection{Underwater Basket
-Weaving}\label{underwater-basket-weaving}
+#### Underwater Basket Weaving
 
-Only two levels of headings should be numbered. Lower level headings
-remain unnumbered; they are formatted as run-in headings.
+Only two levels of headings should be numbered. Lower level headings remain unnumbered; they are formatted as run-in headings.
 
-\paragraph{Underwater Basket Weaving Under Difficult
-Circumstances}\label{underwater-basket-weaving-under-difficult-circumstances}
+##### Underwater Basket Weaving Under Difficult Circumstances
 
-The contribution should contain no more than four levels of headings.
-Table~\ref{tbl-headings} gives a summary of all heading levels.
+The contribution should contain no more than four levels of headings. @tbl-headings gives a summary of all heading levels.
 
-\begin{table}[t]
-
-\caption{\label{tbl-headings}Table captions should be placed above the
-tables.}
-
-\centering{
-
+::: {#tbl-headings tbl-pos='t'}
+```{=latex}
 \begin{tabular}{lll}
@@
 \end{tabular}
+```
 
-}
+Table captions should be placed above the tables.
+:::
 
-\end{table}%
+<!-- | Heading level | Example | Font size and style |
+|------|------|------|
+| Title (centered) | {\Large\bfseries Lecture Notes} | 14 point, bold |
+| 1st-level heading | {\large\bfseries 1 Introduction} | 12 point, bold |
+| 2nd-level heading | {\bfseries 2.1 Printing Area} | 10 point, bold |
+| 3rd-level heading | {\bfseries Run-in Heading in Bold.} Text follows | 10 point, bold |
+| 4th-level heading | {\itshape Lowest Level Heading.} Text follows | 10 point, italic |
 
-\section{Recent Advances from the Mariana
-Trench}\label{recent-advances-from-the-mariana-trench}
+: Table captions should be placed above the
+tables. {#tbl-headings} -->
+
+## Recent Advances from the Mariana Trench
 
@@
 
-\begin{equation}\phantomsection\label{eq-linear}{
+$$
 x + y = z
-}\end{equation}
+$$ {#eq-linear}
+Please try to avoid rasterized images for line-art diagrams and schemas. Whenever possible, use vector graphics instead (see
+@fig-duck).
 
-Please try to avoid rasterized images for line-art diagrams and schemas.
-Whenever possible, use vector graphics instead (see
-Fig.~\ref{fig-duck}).
-
-\begin{figure}[t]
-
-\centering{
-
-\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{example-image-duck.pdf}
-
-}
-
-\caption{\label{fig-duck}A figure caption is always placed below the
-illustration. Please note that short captions are centered, while long
-ones are justified by the macro package automatically.}
-
-\end{figure}%
+![A figure caption is always placed below the illustration.
+Please note that short captions are centered, while long ones are justified by the macro package automatically.](example-image-duck){#fig-duck width=100% fig-pos='t'}
 
@@
 \end{theorem}
-
+<!-- %
+% the environments 'definition', 'lemma', 'proposition', 'corollary',
+% 'remark', and 'example' are defined in the LLNCS documentclass as well.
+% -->
 \begin{proof}
@@
 
-For citations of references, we prefer the use of square brackets and
-consecutive numbers. Citations using labels or the author/year
-convention are also acceptable. The following bibliography provides a
-sample reference list with entries for journal articles
-\cite{ref_article1}, an LNCS chapter \cite{ref_lncs1}, a book
-\cite{ref_book1}, proceedings without editors \cite{ref_proc1}, and a
-homepage \cite{ref_url1}. Multiple citations are grouped
-\cite{ref_article1,ref_lncs1,ref_book1},
-\cite{ref_article1,ref_book1,ref_proc1,ref_url1}.
+<!-- Using Quarto theorem environments does not work, see here: https://github.com/pat-alt/quarto-ecml/issues/1 -->
+<!-- ::: {#thm-sample}
+This is a sample theorem. The run-in heading is set in bold, while the following text appears in italics. Definitions, lemmas, propositions, and corollaries are styled the same way.
+:::
 
-\section{Experiments}\label{experiments}
+::: {.proof}
+Proofs, examples, and remarks have the initial word in italics, while the following text appears in normal font.
+::: -->
 
-\subsection{Experimental Setup}\label{experimental-setup}
+For citations of references, we prefer the use of square brackets and consecutive numbers. Citations using labels or the author/year
+convention are also acceptable.
+The following bibliography provides a sample reference list with entries for journal articles [@ref_article1], an LNCS chapter [@ref_lncs1], a book [@ref_book1], proceedings without editors [@ref_proc1], and a homepage [@ref_url1].
+Multiple citations are grouped [@ref_article1;@ref_lncs1;@ref_book1], [@ref_article1;@ref_book1;@ref_proc1;@ref_url1].
 
-\subsection{Experimental Results}\label{experimental-results}
+## Experiments
 
-\section{Discussion}\label{discussion}
+### Experimental Setup
 
-\section{Conclusion}\label{conclusion}
+### Experimental Results
 
-Of course, authors have complete freedom on how they choose to structure
-their paper. Section headers from Introduction up to and including
-Conclusions are completely up to the discretion of the authors; use
-whichever structure you see fit. Title, Abstract, the credits
-environment, and References, however, are mandatory.
+## Discussion
+
+## Conclusion
+
+Of course, authors have complete freedom on how they choose to structure their paper.
+Section headers from Introduction up to and including Conclusions are completely up to the discretion of the authors; use whichever structure you see fit.
+Title, Abstract, the credits environment, and References, however, are mandatory.
 
@@
 \end{credits}
-
-\bibliographystyle{splncs04}
-\bibliography{bibliography}
-
-\end{document}

### 2025-05-28 Validation_challenges_tree_crown_segmentation/Draft1_PerspectivePaper.tex
add files

% Options for packages loaded elsewhere
% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[runningheads]{llncs}
\usepackage{xcolor}
\usepackage{amsmath,amssymb}
\setcounter{secnumdepth}{2}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[activate=false]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}

\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\newsavebox\pandoc@box
\newcommand*\pandocbounded[1]{% scales image to fit in text height/width
  \sbox\pandoc@box{#1}%
  \Gscale@div\@tempa{\textheight}{\dimexpr\ht\pandoc@box+\dp\pandoc@box\relax}%
  \Gscale@div\@tempb{\linewidth}{\wd\pandoc@box}%
  \ifdim\@tempb\p@<\@tempa\p@\let\@tempa\@tempb\fi% select the smaller of both
  \ifdim\@tempa\p@<\p@\scalebox{\@tempa}{\usebox\pandoc@box}%
  \else\usebox{\pandoc@box}%
  \fi%
}
% Set default figure placement to htbp
\def\fps@figure{htbp}
\makeatother

\setlength{\emergencystretch}{3em} % prevent overfull lines

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage[misc]{ifsym}
\newcommand{\corr}{(\Letter)}
\makeatletter

\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Fig.}
\else
  \newcommand\figurename{Fig.}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter

\makeatother
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Validation challenges in large-scale tree crown segmentations from remote sensing imagery using Deep Learning: a case study in Germany},
  pdfauthor={Taimur Khan, Jasmin Krebs, Nils Nölke, Sharad Kumar Gupta, Jonathan Renkel, Caroline Arnold (order tbd)},
  pdfkeywords={First Keyword, Second Keyword, Another Keyword.},
  colorlinks=true,
  linkcolor={black},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\hypersetup{nolinks=true}

\title{Validation challenges in large-scale tree crown segmentations
from remote sensing imagery using Deep Learning: a case study in
Germany}

\titlerunning{Validation challenges in tree crown segmentations}

\author{Taimur Khan, Jasmin Krebs, Nils Nölke, Sharad Kumar Gupta,
Jonathan Renkel, Caroline Arnold (order tbd)}

\authorrunning{T. Khan et al.}

\institute{}

\begin{document}
\maketitle

\begin{abstract}
Advances in deep learning models have opened up new avenues for computer
vision tasks, like tree crown segmentations from remote sensing images.
As more and more models become available, the challenges of applying the
models to really large spatial scales (beyond the forest/plot level)
become visible. One of these challenges is the validation of the tree
crown detection and segmentation. Imprecise detections/segmentations can
lead to inaccuracies in downstream tasks in forestry and urban planning,
creating a barrier for the application of these models to real-life use
cases. Attributes of training data, e.g tree sizes, phenological stages,
number of polygon vertices, shadows, impact the prediction accuracy of
the segmentation models, hence training data should be chosen with due
diligence. Here we identify some of the key validation challenges based
on the collective experiences of not just training deep learning models,
but also applying the models to very large spatial scales in Germany
(federal state level).

\keywords{First Keyword \and Second Keyword \and Another Keyword.}

\end{abstract}

\section{Introduction}\label{introduction}

Accurately mapping tree crowns over large areas is critical for scaling
local observations to regional and global assessments of forest health,
carbon stocks, and urban greenery. In an era marked by rapid
environmental change, comprehensive mapping efforts are indispensable.
Tree crown segmentation is a cornerstone task for these objectives, as
the size and structure of a tree crown---shaped by species-specific
branching patterns, site conditions, and competition for
light---directly influences primary production. A key metric in this
context is the crown projection area (CPA), defined as the vertical
projection of the crown onto a horizontal plane. Deriving CPA provides
essential insights both at the individual tree level---where it provides
predictions of diameter, volume, and growth rates---and at the stand
level, supporting models of competition and canopy gap dynamics.
However, inaccuracies inherent in the segmentation process, particularly
when applied at large scales, can lead to biased estimations of these
critical tree variables, underscoring the urgent need to address robust
validation challenges in large-scale tree crown segmentation. While
recent advances in deep learning offer unprecedented capabilities for
automated segmentation, their robust validation at scale presents a
critical, unaddressed hurdle.

Among the various tree crown variables, crown spread presents unique
estimation challenges, both in field measurements and, critically, when
derived from remote sensing imagery. Its accurate derivation is not only
susceptible to inaccuracies in the initial segmentation process but also
relies heavily on the methods employed to extract crown spread from
segmented polygons, a process fraught with its own complexities.
Selecting an appropriate crown spread calculation method is crucial,
especially for irregularly shaped tree crown polygons. Even with
optimized calculation methods, direct matches with ground-truth data
proved low (e.g., 32\% in our experience), although a broader acceptable
margin of error (±5 m) captured 89\% of cases. These results immediately
point to difficulties in establishing precise validation metrics.
Furthermore, model performance exhibits size-dependent biases: trees
with small crowns (\textless6 m) are often overestimated, while large
crowns (\textgreater16 m) tend to be underestimated. These discrepancies
underscore the inherent limitations of deep learning segmentation models
at the extremes of the size spectrum. Critically, the quality of
validation data profoundly influences the assessment of crown spread
accuracy. For instance, our experience with the local tree inventory
(`Baumkataster' Halle/Saale), which provides only broad crown spread
ranges (5 m intervals), severely constrains the precision of validation,
highlighting the need for more granular and reliable ground-truth
datasets.

The integration of deep learning into remote sensing has opened new ways
in how we monitor and analyse ecological systems
{[}\cite{zheng2024review}, \cite{zhao2023systematic}{]}. Tree crown
segmentation, a task central to assessing forest health, carbon stocks,
and urban greenery, has especially benefited from advances in
convolutional neural networks (CNNs) and transformer-based
architectures. Approaches leveraging high-resolution imagery have
demonstrated remarkable capabilities in delineating individual tree
crowns, as evidenced by works such as \cite{weinstein2019individual},
\cite{khant2025} and \cite{freudenberg2022individual}, illustrating the
potential for scalable and automated tree mapping. The urgency of
climate change and biodiversity loss further amplifies the importance of
large-scale tree crown segmentation. Forests serve as critical carbon
sinks and biodiversity reservoirs {[}\cite{pan2024enduring}{]}, while
urban trees provide essential ecosystem services
{[}\cite{sharma2024urban}{]}. Large-scale segmentation enables
consistent, detailed monitoring of these vital resources, informing
sustainable forest management, urban planning, and policy development.
The ambition to create national and global inventories, such as those
envisioned in \cite{tolan2024very}, hinges on the reliability of such
foundational segmentation data.

However, while these deep learning models demonstrate impressive
performance in controlled environments or with limited-scale datasets,
their deployment at expansive spatial extents, such as regional or
national scales, exposes a distinct spectrum of challenges that remain
largely unaddressed. Foremost among these is validation---a critical yet
underexplored hurdle in ensuring the reliability and generalizability of
model outputs. As recent studies have shown, the performance of tree
crown segmentation models is highly sensitive to the characteristics of
the training data, such as tree size, species diversity, seasonal
variations, and image quality {[}\cite{moussaid2021tree};
\cite{cong2022citrus}{]}. These factors, when coupled with the
complexities of real-world landscapes, underscore the need for robust
and scalable validation frameworks.

This Perspective builds on our extensive experience in applying deep
learning models to large-scale tree crown segmentation across Germany,
addressing federal state-level landscapes. By synthesizing insights from
model training, deployment, and validation, we highlight the nuanced
challenges associated with assessing segmentation accuracy over
expansive and heterogeneous terrains. These challenges are not mere
technicalities---they bear significant implications for downstream
applications in forestry, conservation, and urban planning. Through this
work, we aim to not only highlight these critical validation hurdles but
also to foster a dialogue on the methodological innovations required to
ensure that deep learning advancements in remote sensing translate into
actionable insights at the scales demanded by global environmental
challenges.

\section{Case Study: Tree Crown Segmentation in Saxony and
Saxony-Anhalt,
Germany}\label{case-study-tree-crown-segmentation-in-saxony-and-saxony-anhalt-germany}

In this case study, the DeepTrees model {[}\cite{khan2025}{]} was
applied in a one-shot prediction approach using pretrained model weights
provided by DeepTrees, applied to high-resolution multispectral digital
orthophoto imagery (DOP20) covering the German federal states of Saxony
(SN) and Saxony-Anhalt (ST). Sachsen, covering approximately 18,450 km²,
and Saxony-Anhalt, spanning around 20,452 km², represent diverse
ecological and urban landscapes, ideal for assessing large-scale
segmentation model performance.

DOP20 imagery provides detailed spatial information at 20 cm resolution
per pixel, capturing visual and near infrared spectral regions essential
for accurate tree crown delineation. Using these pretrained weights, the
DeepTrees model identified approximately 218.7 million individual tree
crowns---137.3 million in Saxony and 81.4 million in Saxony-Anhalt. The
resulting segmentation dataset has been made publicly available on
Zenodo (https://zenodo.org/record/exampledoi), facilitating further
research and validation.

This comprehensive segmentation highlights substantial regional
variability in tree distribution, reflecting ecological, topographical,
and land-use gradients. The case study thus forms a practical foundation
for addressing the validation challenges detailed subsequently,
emphasizing the necessity for robust validation frameworks to ensure
reliable ecological monitoring and informed decision-making based on
large-scale automated analyses.

\emph{Figure 1: 6 panels (1 for each state). From left to right
-\textgreater{} 1) tiles with date classification 2) Land use (CORINE) +
isolines. 3) the tree segmentations.}

\section{Challenges}\label{challenges}

Validating large-scale tree crown segmentation models reveals a web of
interrelated challenges that go beyond those encountered in small,
controlled datasets. Seasonal phenology is a moving target: the same
forest can look drastically different between leaf-on summer imagery and
leaf-off winter scenes. Models trained on one season often struggle in
another, yielding inconsistent segmentation accuracy across the year.
For example, a canopy delineation that performs well on lush summer
foliage may under-segment sparse autumn crowns or miss bare branches in
winter. Such season-driven variability not only degrades model
performance but also complicates validation -- a one-shot model might
appear accurate in one season and fail in the next, raising questions
about how and when accuracy should be assessed. Incorporating
multi-season data during both training and validation is essential, as
phenological dynamics have been shown to strongly influence model
generalizability {[}\cite{moussaid2021tree}; \cite{cong2022citrus}{]}.

Spatial heterogeneity of landscapes, including terrain and illumination
differences, poses another major hurdle for both segmentation and its
validation. An algorithm that segments tree crowns flawlessly in a flat,
well-lit park may stumble in a shadow-drenched valley or on a steep
mountainside. Variations in ground elevation and slope alter the angle
of solar illumination, leading to uneven lighting and shadows that can
confuse models. In mountainous or rugged terrains, trees on north-facing
slopes might appear darker or partially occluded compared to those on
south-facing slopes with direct sun, even if they are the same species
and healthy. Such effects result in site-specific performance: models
often need fine-tuning or can suffer accuracy drops when moved to new
topographies or sensor angles. Weinstein et al.~(2020) observed this
kind of cross-site performance gap, where a tree detection model trained
in one region underperformed when applied to a different region's
imagery without adaptation, underscoring how terrain and context
influence outcomes {[}\cite{weinstein2020cross}{]}. For validation, this
means that accuracy estimates from one area may not transfer to another
-- a critical issue when assessments at national or global scales are
required.

A further fundamental challenge lies in the scarcity of accurate
ground-truth data at scale. Reliable validation hinges on high-quality
reference data (the ``ground truth''), yet collecting detailed crown
delineations over large regions is logistically difficult and expensive.
Field surveys can map individual trees with great precision (e.g.,
measuring trunks and canopy spread on the ground), but doing this over
thousands of square kilometers is infeasible. Aerial and satellite
imagery provide broader coverage for reference data, but even these
often lack the resolution or fidelity to unequivocally label each tree
crown for validation purposes. UAV (drone) campaigns can bridge the gap
by capturing very high-resolution images or LiDAR of sample areas, but
they are limited in flight range and still require extensive human
annotation to turn imagery into usable ground truth. The net result is a
mismatch of scales: our models aspire to map every tree across entire
countries, but our ground truth typically covers only small plots or
scattered samples {[}\^{}{]}. This mismatch means that validating a
``wall-to-wall'' tree map often involves extrapolating from a tiny
fraction of ground-referenced trees, introducing uncertainty. Moreover,
ground-reference datasets may not capture the full diversity of
conditions (species, canopy shapes, management regimes, etc.) present in
the larger mapping area, biasing the validation. Expanding ground-truth
collection -- through automated methods or crowdsourcing -- is thus not
just a recommendation but a necessity to overcome this validation
bottleneck (as we discuss later).

Compounding the issue of limited data is the inconsistency in reference
annotations and evaluation metrics. Even when ground-reference data
exist, their format can differ -- sometimes reference trees are marked
by a single GPS point (e.g., trunk location), sometimes by a hand-drawn
polygon outlining the crown. This creates a challenge in validation: how
do we decide if a predicted crown polygon ``matches'' a ground-truth
point, or how to handle cases where one field-mapped tree corresponds to
multiple overlapping crown segments in the image?(Figure ref)
Conversely, field crews might delineate a broad canopy as one crown
while an automated model splits it into two segments (or vice versa),
especially in dense stands where crowns merge. These ambiguities in
one-to-one correspondence make it hard to define what a ``correct''
segmentation is. Traditional pixel-wise accuracy metrics like
Intersection-over-Union (IoU) treat segmentation purely as an image
overlap problem, which may not reflect the ecological reality of
counting individual trees. IoU penalizes differences in shape or area
but doesn't account for whether the count of tree objects is correct. In
an extreme case, a model could slightly over-segment every tree
(splitting each true crown into two smaller polygons) and still achieve
a reasonable IoU, despite doubling the perceived tree count -- a
significant error for applications. On the other hand, object-centric
metrics such as panoptic segmentation quality attempt to consider both
detection and delineation of objects
{[}\cite{kirillov2019panopticsegmentation}{]}. Panoptic metrics combine
aspects of object detection (was each tree detected?) with segmentation
quality (was each crown correctly outlined?), which can be more
appropriate for tree mapping. However, even these require well-defined
ground-truth objects to compare against. When the ground truth itself is
inconsistent (e.g., how to count a clumped cluster of stems with
overlapping crowns), validation metrics struggle to fully capture model
performance. The choice of evaluation metric thus becomes non-trivial:
depending on whether one prioritizes exact crown shape, tree count, or
canopy cover, the ``best'' metric may differ. Establishing consensus on
evaluation protocols is part of the challenge -- without it, different
studies may report accuracy in incompatible ways.

There is also the issue of scale and resolution in validation reporting.
A model's accuracy can appear to vary depending on the spatial scale at
which it's evaluated. For instance, a segmentation model might achieve
high overall accuracy when averaged over an entire region, yet if one
zooms into a small test area (say a single city park or forest stand),
the error rate might be much higher or lower. This can happen if errors
are not evenly distributed: the model could perform very well in one
type of landscape (e.g., neat urban street trees) and poorly in another
(dense natural forest), and a coarse regional average could mask these
extremes. Consequently, a user working on a local conservation project
might experience worse performance than the ``headline'' accuracy
suggests, because that headline number was diluted by many easier cases
elsewhere. Ensuring that validation is robust across scales is tricky --
one must balance broad coverage with local detail. It calls for
multi-scale validation approaches, where accuracy is reported at
multiple grain sizes or stratified by landscape type. Highlighting this,
one could imagine a figure plotting model accuracy as a function of
spatial extent or across different habitat types, illustrating how
performance can drop off in specific challenging subsets despite looking
good overall. This emphasizes that heterogeneous performance is itself a
challenge to acknowledge in validation: we need methods to detect where
and why a model fails, not just an aggregate score.

To summarize, validating large-scale tree segmentation models is far
from a straightforward task. Seasonal changes, diverse terrain and
illumination, limited and inconsistent ground truth, ambiguous
evaluation criteria, and scale-dependent performance all intertwine to
create a demanding setting. These challenges are not just academic --
they directly impact how much trust we can place in AI-generated tree
maps for real-world decisions in forestry, ecology, and urban planning.
Recognizing these pain points is the first step; the next is devising
strategies to overcome them. We now turn to several recommendations
aimed at improving both models and validation practices, with an eye
toward bridging the gap between controlled experiments and the
complexity of continental-scale deployments.

\emph{Table 1: Potential validation challenges and corresponding
mitigation strategies. This table could list key challenges in
validating large-scale tree segmentations alongside proposed approaches
to address them. For example: (Challenge) Phenological variation between
seasons -- (Mitigation) use multi-temporal training data and
season-specific validation sets; (Challenge) Heterogeneous terrain and
shadowing -- (Mitigation) integrate digital elevation models and
terrain-aware modeling; (Challenge) Scarcity of ground-truth annotations
-- (Mitigation) leverage semi-automated labeling, crowdsourcing, and
active learning to expand validation data; (Challenge) Ambiguity in
crown delineation -- (Mitigation) adopt object-based accuracy metrics
and consensus protocols for what counts as a single tree; (Challenge)
Scale-dependent performance -- (Mitigation) evaluate models at multiple
spatial scales and stratify results by landscape type. This overview
would visually reinforce the narrative that each challenge has an
identifiable path forward, linking the ``Challenges'' and
``Recommendations'' sections.}

\emph{!{[}{]}{[}image1{]}}\\
\emph{Figure 2: Incomplete}

\section{Recommendations}\label{recommendations}

Addressing the above challenges requires a multi-pronged approach,
combining improvements in modeling techniques with innovations in
validation methodology and data collection. We outline here several
complementary strategies that, together, can bolster the reliability of
large-scale tree segmentation efforts. These recommendations emphasize
building more robust models through advanced training paradigms, as well
as creating better frameworks to evaluate and support those models in
real-world conditions. The overarching goal is to ensure that deep
learning advancements in tree mapping translate into trustworthy,
actionable insights at regional to global scales.

\textbf{Leverage self-supervised learning for robust representations.} A
key step towards more generalizable segmentation models is tapping into
unannotated data via self-supervised learning (SSL). Unlike traditional
supervised training which is bottlenecked by limited labeled examples,
SSL allows models to learn from the abundant pool of unlabeled remote
sensing images -- for example, by predicting missing pieces of an image
or distinguishing augmented views of the same scene. By pre-training a
model on large geospatial datasets without any human-provided labels,
the model can absorb intrinsic patterns of the landscape: textures,
shapes, seasonal changes, and other context that are common across
images. These rich foundational representations can then be fine-tuned
for tree crown segmentation with far fewer annotated samples than would
otherwise be needed. Recent efforts like PhilEO Bench demonstrate the
promise of this approach, evaluating geospatial foundation models that
were pre-trained using SSL techniques {[}\cite{fibaek2024phileo}{]}. The
results show improved performance on a range of tasks (e.g.~building
footprint extraction, road mapping) after such pre-training, compared to
models trained from scratch. In the context of trees, a model with an
SSL-pretrained backbone may already ``know'' about basic vegetation
structures, shadows, and seasonal appearances, making it more adept at
delineating crowns under varied conditions. For instance, a foundation
model trained on year-round satellite imagery might implicitly
understand the difference between a leafless oak in winter and the same
oak in summer, and thus require only a light fine-tuning to accurately
segment each. Mendieta et al.~(2023) take this further by continually
pre-training on new data distributions (a form of continual learning),
which helped build a Geospatial Foundation Model (GFM) that excelled
across multiple remote sensing tasks {[}\cite{mendieta2023gfm}{]}. Such
continual SSL training could allow segmentation models to keep improving
as more unlabeled data (e.g., new satellite images over time or from new
regions) become available, staying up-to-date with changes in landscapes
and sensor characteristics. Exploiting SSL is a powerful way to tackle
the twin issues of limited labels and dataset bias, yielding a model
that is better equipped for the diverse scenarios encountered in
large-scale mapping.

\textbf{Integrate multi-view and multi-temporal data for consistency.}
Another promising avenue is to train and evaluate models with multi-view
inputs -- that is, imagery of the same trees captured from different
angles, sensors, or times. Multi-view here can mean different things:
multi-angle (such as combining nadir and oblique aerial images),
multi-platform (satellite imagery plus drone photos), or multi-temporal
(images taken in different seasons or years). By exposing models to such
spatiotemporal diversity, we can help them learn invariances that make
segmentation more reliable. For example, a tree crown seen from directly
above might have a certain shape, but from a side angle the outline
might blend with neighbors; a model trained with both perspectives could
learn to robustly identify the crown in either view. Likewise, combining
leaf-on and leaf-off images of the same forest during training can force
a model to rely on structural features (branches, trunk hints, relative
spacing) in addition to just greenness, thereby improving its ability to
generalize across seasons. Self-supervised methods are particularly
well-suited to capitalize on multi-view data because they can be set up
to encourage the model to produce similar latent representations for
different views of the same object. Techniques such as masked
autoencoders or contrastive learning can be used on multi-view datasets
to make a model predict or align one view from another, without any
manual labels needed. Such approaches have already shown success: for
instance, researchers have used masked image modeling and contrastive
SSL on multi-view satellite imagery to boost performance on segmentation
and detection tasks, essentially teaching the model that ``these two
different-looking images actually contain the same trees''
{[}\cite{mukkavilli2023prithvi}; \cite{fibaek2024phileo}{]}. By training
with multi-view consistency, the model becomes more robust to viewpoint
and temporal changes, which directly addresses the phenology and
illumination challenges discussed earlier. Importantly, this not only
helps the model's predictions but also strengthens validation, because a
model that is consistent across views makes it easier to compare
predicted and true crowns even when the reference data comes from a
slightly different perspective or date. In practice, one could imagine a
validation scheme where the agreement of a model's output between
leaf-on and leaf-off images serves as an indicator of reliability: large
discrepancies might flag areas for further human inspection. Overall,
multi-view and multi-temporal training imbue models with a form of
contextual intelligence about the 3D and time-varying nature of trees,
making segmentation outcomes more stable across the real-world
variability that large-scale applications inevitably encompass.

\textbf{Incorporate terrain information into the modeling pipeline}. As
noted, uneven terrain can cause substantial variability in how trees
appear in images, so bringing explicit knowledge of terrain into the
segmentation process is a logical remedy. One recommendation is to fuse
digital elevation models (DEMs) or LiDAR-derived terrain data with the
imagery during model training. This could be as simple as providing the
model with an extra channel of input encoding elevation/slope, or as
complex as designing the model to separately process terrain context.
Self-supervised pre-training can be extended here too: recent work on
multisensor geospatial foundation models has shown that including
elevation data in SSL (e.g., tasking the model to distinguish imagery of
flat vs.~mountainous areas) yields better feature representations for
downstream tasks {[}\cite{han2024msGFM}{]}. By differentiating between
bare earth and above-ground structures in pre-training, the model learns
to discount illumination differences that are purely due to slope and
aspect, focusing instead on actual objects like trees. In a tree
segmentation scenario, a terrain-informed model could recognize that a
dark region in an image is a shaded hillside rather than a non-existent
gap in canopy cover, or that an elongated shape on a steep slope is
still a single tree crown albeit skewed by perspective. Incorporating
terrain data directly addresses the challenge of spatial heterogeneity:
it provides a reference frame to normalize out some variability. This
can improve validation as well -- for example, error analysis can be
stratified by terrain class to ensure a model works not just on average,
but on hilltops and valleys alike. We recommend that future segmentation
models, especially for large regions with varied topography, adopt
terrain-aware training strategies. Even if a full DEM is not available
everywhere, approximating slope from the imagery or using coarse global
elevation data could still offer benefits. Ultimately, bridging the gap
between pixel appearance and real-world topography makes the model's
understanding more physical and generalizable, reducing surprises when
it's deployed on a new landscape.

\textbf{Focus model attention on domain-specific features.} Beyond data
augmentation and multi-source inputs, improvements in the model
architecture and training objectives themselves can yield more reliable
segmentation. One intriguing direction is the use of feature-guided
masked autoencoders or similar techniques that encourage the model to
learn high-level features rather than getting bogged down in pixel-level
noise. In remote sensing, not all pixels are equal -- the spectral
signature of a healthy tree canopy, for instance, is characterized by
certain reflectance patterns (like high near-infrared reflectance for
foliage), and textural cues can differentiate a tree from grass. A
vanilla model might or might not latch onto these subtle cues. However,
a feature-guided approach explicitly trains the model to reconstruct or
predict meaningful feature representations (such as vegetation indices
or edge maps) instead of raw pixel intensity. For example, a recent
approach called FG-MAE (Feature Guided Masked Autoencoder) masks out
parts of an image and tasks the model with predicting domain-relevant
features (like a NDVI -- Normalized Difference Vegetation Index image,
or other engineered representations) for the masked region rather than
the raw pixels. This forces the model to infer what type of object
should be there, not just to copy textures, thereby learning a more
semantic understanding of the scene {[}\cite{mukkavilli2023prithvi}{]}.
Applying this idea to tree segmentation, we could pre-train models to
predict features that highlight vegetation structure (perhaps oriented
gradients that capture tree crown edges, or canopy height estimations
from stereo images) so that the model's internal representations become
highly attuned to ``tree-ness.'' When fine-tuned for segmentation, such
a model may be better at delineating tangled canopies or differentiating
trees from confusing background elements, because it has learned to
focus on the attributes that define a tree in imagery. Early studies in
multispectral and SAR domains have found that this approach yields
improved performance in complex environments
{[}\cite{mukkavilli2023prithvi}{]}. We recommend integrating
feature-guidance in training for large-scale tree mapping, especially in
areas with complex backgrounds (e.g., urban environments where trees
mingle with buildings). Not only does this likely boost accuracy, it
could also produce more interpretable model outputs or uncertainties --
a model that knows what features it's looking for might provide
reasoning (explicit or implicit) for its segmentation, which in turn
aids validation and error diagnosis.

\textbf{Expand validation beyond pixel agreement -- use ecological
consistency checks.} Traditional validation of segmentation focuses on
geometric overlap with ground truth shapes, but for tree mapping we can
also exploit well-established ecological relationships as an additional
form of validation. Trees have allometric relationships -- mathematical
links between dimensions like trunk diameter, height, crown diameter,
and biomass. These relationships have been measured in forestry for
decades. For example, a tree of a given height typically has a crown of
roughly proportional diameter, and there are known bounds on how big a
crown can get for a given trunk size. We propose using such allometric
equations as a sanity check for segmentation outputs. If a model's
predicted tree crowns violate basic allometry (say, a tiny tree height
but a huge crown width, or a cluster of crowns whose total basal area
implies an impossible biomass for that area), it might indicate errors
in the segmentation or missing trees. One could quantitatively compare
the distribution of predicted crown sizes and tree heights (if height
data or estimates are available) against expected distributions from
field data. Significant deviations could flag problems: for instance, if
the model frequently delineates very large crowns that in reality would
correspond to 80-meter tall trees (which don't exist in the region),
those are likely over-segmentation artifacts. Conversely, if predicted
crowns are all very small in an old-growth forest where trees should be
large, the model might be under-segmenting (splitting one crown into
many). Researchers have indeed used this kind of approach in related
contexts; for example, Song et al.~(2023) employed Gaussian process
regression on tree height to estimate biomass, demonstrating how linking
remote sensing outputs to allometric models can validate whether the
outputs make ecological sense {[}\cite{song2023biomass}{]}. In practice,
implementing allometric checks means bringing in additional data or
models (e.g., a LiDAR-derived tree height map, or species-specific
allometric formulas from forestry literature) and cross-verifying the
plausibility of the AI-generated tree map. This recommendation shifts
validation from a purely computer-vision perspective to an
application-oriented perspective: after all, if the ultimate goal is to
use these maps for carbon accounting, biodiversity, or forestry, then
passing an ecological reality check is as important as scoring well on
IoU.

\textbf{Establish community benchmarks and evaluation frameworks.} The
field of geospatial AI is recognizing the value of standard benchmarks
-- datasets and metrics on which different methods can be compared in a
reproducible way. For computer vision in general, ImageNet and COCO
served this role, and now geospatial analogues are emerging. Efforts
like GEO-Bench and PANGAEA have started assembling diverse geospatial
tasks into comprehensive evaluation suites, underscoring the importance
of benchmarking models across varying conditions
{[}\cite{marsocci2024pangaea}{]}. We recommend developing similar
benchmark datasets and challenges specifically for tree crown
segmentation and mapping. These benchmarks should cover a wide range of
landscapes (different forest types, urban trees, plantations, etc.),
seasons, and remote sensing data sources, reflecting the real-world
diversity highlighted in our Challenges section. Importantly, a common
benchmark would promote consensus on validation metrics -- if the
community agrees to evaluate on, say, a mix of IoU for segmentation
quality and perhaps a detection metric for counting accuracy, it would
standardize how results are reported. This comparability accelerates
progress: researchers can pinpoint which innovations truly lead to
better performance and robustness. Moreover, a shared evaluation
framework can include protocols for multi-scale assessment (e.g.,
requiring methods to report not just overall accuracy but also
performance on hard subsets like ``dense forest canopy'' vs ``isolated
trees'') and encourage inclusion of auxiliary checks like the allometric
consistency mentioned above. In short, a benchmarking initiative for
large-scale tree segmentation would provide an invaluable feedback loop,
where challenges discovered by one team (e.g., failure cases in a
certain region or condition) become part of the test set that everyone
then tries to solve. This kind of iterative improvement cycle has been
crucial in other domains of AI. We envision something like an annual
challenge where models compete on a standard large-scale tree mapping
task -- pushing methods to be not only accurate but also general and
validation-friendly. By establishing these community standards, we move
toward a future where claims of model performance are transparent and
believable, because they've been vetted on a broad, agreed-upon spectrum
of scenarios.

\textbf{Innovate in ground-truth data collection and labeling.} Lastly,
none of the above model improvements obviates the need for better and
more plentiful validation data. We therefore recommend parallel efforts
to enhance ground-truth collection through automation and crowdsourcing.
On the automation side, advances in high-resolution imaging and onboard
AI mean that drones or small aircraft could be deployed to automatically
detect and delineate trees in sample areas, providing semi-automated
annotations that experts only need to lightly verify. For instance, a
drone flying over a forest could use its own simpler model to suggest
crown boundaries, which are then corrected by a human specialist --
vastly speeding up the annotation process compared to drawing polygons
from scratch. Similarly, leveraging crowdsourcing and citizen science
could dramatically expand validation datasets: non-experts can be asked
to label trees on accessible platforms (especially for easy tasks like
clicking the center of a visible tree), and with enough redundancy and
quality control, these contributions can rival expert labels. Projects
for mapping trees in cities (e.g., through apps that engage the public
to identify street trees) hint at this potential. Of course, when using
crowd data, one must account for variability in skill and ensure
rigorous cleaning and validation of the contributed labels -- but as a
complementary source of truth data, it could fill gaps in areas
professionals have not covered. Another approach is active learning,
where the model itself guides where more data is needed: the model can
flag areas or examples where it is most uncertain or makes conflicting
predictions (say, an area of weirdly segmented crowns), and those areas
would be prioritized for human labeling. This way, instead of randomly
sampling locations for ground truth, we focus effort on the most
informative samples -- those likely to teach the model something new or
reveal a blind spot. Active learning strategies have been shown to
greatly reduce the amount of data needed for training in other remote
sensing tasks by smartly selecting the right samples to label
{[}\^{}{]}. Integrating an active learning loop into large-scale
validation means we continually refine the ground truth in the areas
that matter most, thereby improving the model in a targeted fashion.
Finally, we should not overlook the role of policy and open data
initiatives: governments and organizations conducting tree inventories
or LiDAR scans should be encouraged to share these as open benchmarks.
Even if such datasets are not perfectly aligned with satellite imagery,
they can often be matched or used to validate parts of a map. In
summary, closing the validation data gap will require creativity --
using machines to help label, using the crowd to scale up, and using the
model's own intelligence to guide where to look next. These efforts
complement the technical recommendations above: better models make use
of the new data more effectively, and better data enables more powerful
models.

By pursuing these recommendations in concert, the field can make
significant strides toward accurate and trustworthy large-scale tree
mapping. What we envision is an end-to-end pipeline where robust models
(forged through self-supervised, multi-view, terrain-aware training)
produce segmentation maps that are continuously checked against both
classical metrics and real-world plausibility (via benchmarks and
allometric/field validations), and where the feedback from validation
drives further improvement of the models through active learning and
expanded training data.

\bibliographystyle{splncs04}
\bibliography{bibliography}

\end{document}

### 2025-05-28 Validation_challenges_tree_crown_segmentation/Draft1_PerspectivePaper.qmd
add qmd file

--- 2025-05-28 Validation_challenges_tree_crown_segmentation/Draft1_PerspectivePaper.tex
@@
-% Options for packages loaded elsewhere
-% Options for packages loaded elsewhere
-\PassOptionsToPackage{unicode}{hyperref}
-\PassOptionsToPackage{hyphens}{url}
-\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
-%
-\documentclass[runningheads]{llncs}
-\usepackage{xcolor}
-\usepackage{amsmath,amssymb}
-\setcounter{secnumdepth}{2}
-\usepackage{iftex}
-\ifPDFTeX
-  \usepackage[T1]{fontenc}
-  \usepackage[utf8]{inputenc}
-  \usepackage{textcomp} % provide euro and other symbols
-\else % if luatex or xetex
-  \usepackage{unicode-math} % this also loads fontspec
-  \defaultfontfeatures{Scale=MatchLowercase}
-  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
-\fi
-\usepackage{lmodern}
-\ifPDFTeX\else
-  % xetex/luatex font selection
-\fi
-% Use upquote if available, for straight quotes in verbatim environments
-\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
-\IfFileExists{microtype.sty}{% use microtype if available
-  \usepackage[activate=false]{microtype}
-  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
-}{}
+---
+title: "Validation challenges in large-scale tree crown segmentations from remote sensing imagery using Deep Learning: a case study in Germany"
+title-running: Validation challenges in tree crown segmentations
+format:
+  quarto-ecml-pdf:
+    keep-tex: true
+  quarto-ecml-html: default
+author: Taimur Khan, Jasmin Krebs, Nils Nölke, Sharad Kumar Gupta, Jonathan Renkel, Caroline Arnold (order tbd)
+author-running: T. Khan et al.
+abstract: |
+  Advances in deep learning models have opened up new avenues for computer vision tasks, like tree crown segmentations from remote sensing images.
+As more and more models become available, the challenges of applying the models to really large spatial scales (beyond the forest/plot level) become visible.
+One of these challenges is the validation of the tree crown detection and segmentation.
+Imprecise detections/segmentations can lead to inaccuracies in downstream tasks in forestry and urban planning, creating a barrier for the application of these models to real-life use cases.
+Attributes of training data, e.g tree sizes, phenological stages, number of polygon vertices, shadows, impact the prediction accuracy of the segmentation models, hence training data should be chosen with due diligence.
+Here we identify some of the key validation challenges based on the collective experiences of not just training deep learning models, but also applying the models to very large spatial scales in Germany (federal state level).
+keywords: [First Keyword, Second Keyword, Another Keyword.]
+---
 
-\usepackage{longtable,booktabs,array}
-\usepackage{calc} % for calculating minipage widths
-% Correct order of tables after \paragraph or \subparagraph
-\usepackage{etoolbox}
-\makeatletter
-\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
-\makeatother
-% Allow footnotes in longtable head/foot
-\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
-\makesavenoteenv{longtable}
-\usepackage{graphicx}
-\makeatletter
-\newsavebox\pandoc@box
-\newcommand*\pandocbounded[1]{% scales image to fit in text height/width
-  \sbox\pandoc@box{#1}%
-  \Gscale@div\@tempa{\textheight}{\dimexpr\ht\pandoc@box+\dp\pandoc@box\relax}%
-  \Gscale@div\@tempb{\linewidth}{\wd\pandoc@box}%
-  \ifdim\@tempb\p@<\@tempa\p@\let\@tempa\@tempb\fi% select the smaller of both
-  \ifdim\@tempa\p@<\p@\scalebox{\@tempa}{\usebox\pandoc@box}%
-  \else\usebox{\pandoc@box}%
-  \fi%
-}
-% Set default figure placement to htbp
-\def\fps@figure{htbp}
-\makeatother
+## Introduction
 
-\setlength{\emergencystretch}{3em} % prevent overfull lines
+Accurately mapping tree crowns over large areas is critical for scaling local observations to regional and global assessments of forest health, carbon stocks, and urban greenery.
+In an era marked by rapid environmental change, comprehensive mapping efforts are indispensable.
+Tree crown segmentation is a cornerstone task for these objectives, as the size and structure of a tree crown—shaped by species-specific branching patterns, site conditions, and competition for light—directly influences primary production.
+A key metric in this context is the crown projection area (CPA), defined as the vertical projection of the crown onto a horizontal plane.
+Deriving CPA provides essential insights both at the individual tree level—where it provides predictions of diameter, volume, and growth rates—and at the stand level, supporting models of competition and canopy gap dynamics.
+However, inaccuracies inherent in the segmentation process, particularly when applied at large scales, can lead to biased estimations of these critical tree variables, underscoring the urgent need to address robust validation challenges in large-scale tree crown segmentation.
+While recent advances in deep learning offer unprecedented capabilities for automated segmentation, their robust validation at scale presents a critical, unaddressed hurdle.
 
-\providecommand{\tightlist}{%
-  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
+Among the various tree crown variables, crown spread presents unique estimation challenges, both in field measurements and, critically, when derived from remote sensing imagery.
+Its accurate derivation is not only susceptible to inaccuracies in the initial segmentation process but also relies heavily on the methods employed to extract crown spread from segmented polygons, a process fraught with its own complexities.
+Selecting an appropriate crown spread calculation method is crucial, especially for irregularly shaped tree crown polygons.
+Even with optimized calculation methods, direct matches with ground-truth data proved low (e.g., 32% in our experience), although a broader acceptable margin of error (±5 m) captured 89% of cases.
+These results immediately point to difficulties in establishing precise validation metrics.
+Furthermore, model performance exhibits size-dependent biases: trees with small crowns (\<6 m) are often overestimated, while large crowns (\>16 m) tend to be underestimated.
+These discrepancies underscore the inherent limitations of deep learning segmentation models at the extremes of the size spectrum.
+Critically, the quality of validation data profoundly influences the assessment of crown spread accuracy.
+For instance, our experience with the local tree inventory ('Baumkataster' Halle/Saale), which provides only broad crown spread ranges (5 m intervals), severely constrains the precision of validation, highlighting the need for more granular and reliable ground-truth datasets.
 
-\usepackage[T1]{fontenc}
-\usepackage{graphicx}
-\usepackage{booktabs}
-\usepackage[misc]{ifsym}
-\newcommand{\corr}{(\Letter)}
-\makeatletter
+The integration of deep learning into remote sensing has opened new ways in how we monitor and analyse ecological systems \[@zheng2024review, @zhao2023systematic\].
+Tree crown segmentation, a task central to assessing forest health, carbon stocks, and urban greenery, has especially benefited from advances in convolutional neural networks (CNNs) and transformer-based architectures.
+Approaches leveraging high-resolution imagery have demonstrated remarkable capabilities in delineating individual tree crowns, as evidenced by works such as @weinstein2019individual, @khant2025 and @freudenberg2022individual, illustrating the potential for scalable and automated tree mapping.
+The urgency of climate change and biodiversity loss further amplifies the importance of large-scale tree crown segmentation.
+Forests serve as critical carbon sinks and biodiversity reservoirs \[@pan2024enduring\], while urban trees provide essential ecosystem services \[@sharma2024urban\].
+Large-scale segmentation enables consistent, detailed monitoring of these vital resources, informing sustainable forest management, urban planning, and policy development.
+The ambition to create national and global inventories, such as those envisioned in @tolan2024very, hinges on the reliability of such foundational segmentation data.
 
-\AtBeginDocument{%
-\ifdefined\contentsname
-  \renewcommand*\contentsname{Table of contents}
-\else
-  \newcommand\contentsname{Table of contents}
-\fi
-\ifdefined\listfigurename
-  \renewcommand*\listfigurename{List of Figures}
-\else
-  \newcommand\listfigurename{List of Figures}
-\fi
-\ifdefined\listtablename
-  \renewcommand*\listtablename{List of Tables}
-\else
-  \newcommand\listtablename{List of Tables}
-\fi
-\ifdefined\figurename
-  \renewcommand*\figurename{Fig.}
-\else
-  \newcommand\figurename{Fig.}
-\fi
-\ifdefined\tablename
-  \renewcommand*\tablename{Table}
-\else
-  \newcommand\tablename{Table}
-\fi
-}
-\@ifpackageloaded{float}{}{\usepackage{float}}
-\floatstyle{ruled}
-\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
-\floatname{codelisting}{Listing}
-\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
-\makeatother
-\makeatletter
-\makeatother
-\makeatletter
+However, while these deep learning models demonstrate impressive performance in controlled environments or with limited-scale datasets, their deployment at expansive spatial extents, such as regional or national scales, exposes a distinct spectrum of challenges that remain largely unaddressed.
+Foremost among these is validation—a critical yet underexplored hurdle in ensuring the reliability and generalizability of model outputs.
+As recent studies have shown, the performance of tree crown segmentation models is highly sensitive to the characteristics of the training data, such as tree size, species diversity, seasonal variations, and image quality \[@moussaid2021tree; @cong2022citrus\].
+These factors, when coupled with the complexities of real-world landscapes, underscore the need for robust and scalable validation frameworks.
 
-\makeatother
-\usepackage{bookmark}
-\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
-\urlstyle{same}
-\hypersetup{
-  pdftitle={Validation challenges in large-scale tree crown segmentations from remote sensing imagery using Deep Learning: a case study in Germany},
-  pdfauthor={Taimur Khan, Jasmin Krebs, Nils Nölke, Sharad Kumar Gupta, Jonathan Renkel, Caroline Arnold (order tbd)},
-  pdfkeywords={First Keyword, Second Keyword, Another Keyword.},
-  colorlinks=true,
-  linkcolor={black},
-  filecolor={Maroon},
-  citecolor={Blue},
-  urlcolor={Blue},
-  pdfcreator={LaTeX via pandoc}}
+This Perspective builds on our extensive experience in applying deep learning models to large-scale tree crown segmentation across Germany, addressing federal state-level landscapes.
+By synthesizing insights from model training, deployment, and validation, we highlight the nuanced challenges associated with assessing segmentation accuracy over expansive and heterogeneous terrains.
+These challenges are not mere technicalities—they bear significant implications for downstream applications in forestry, conservation, and urban planning.
+Through this work, we aim to not only highlight these critical validation hurdles but also to foster a dialogue on the methodological innovations required to ensure that deep learning advancements in remote sensing translate into actionable insights at the scales demanded by global environmental challenges.
 
-\hypersetup{nolinks=true}
+## Case Study: Tree Crown Segmentation in Saxony and Saxony-Anhalt, Germany
 
-\title{Validation challenges in large-scale tree crown segmentations
-from remote sensing imagery using Deep Learning: a case study in
-Germany}
+In this case study, the DeepTrees model \[@khan2025\] was applied in a one-shot prediction approach using pretrained model weights provided by DeepTrees, applied to high-resolution multispectral digital orthophoto imagery (DOP20) covering the German federal states of Saxony (SN) and Saxony-Anhalt (ST).
+Sachsen, covering approximately 18,450 km², and Saxony-Anhalt, spanning around 20,452 km², represent diverse ecological and urban landscapes, ideal for assessing large-scale segmentation model performance.
 
-\titlerunning{Validation challenges in tree crown segmentations}
+DOP20 imagery provides detailed spatial information at 20 cm resolution per pixel, capturing visual and near infrared spectral regions essential for accurate tree crown delineation.
+Using these pretrained weights, the DeepTrees model identified approximately 218.7 million individual tree crowns—137.3 million in Saxony and 81.4 million in Saxony-Anhalt.
+The resulting segmentation dataset has been made publicly available on Zenodo (https://zenodo.org/record/exampledoi), facilitating further research and validation.
 
-\author{Taimur Khan, Jasmin Krebs, Nils Nölke, Sharad Kumar Gupta,
-Jonathan Renkel, Caroline Arnold (order tbd)}
+This comprehensive segmentation highlights substantial regional variability in tree distribution, reflecting ecological, topographical, and land-use gradients.
+The case study thus forms a practical foundation for addressing the validation challenges detailed subsequently, emphasizing the necessity for robust validation frameworks to ensure reliable ecological monitoring and informed decision-making based on large-scale automated analyses.
 
-\authorrunning{T. Khan et al.}
+*Figure 1: 6 panels (1 for each state). From left to right \-\> 1\) tiles with date classification 2\) Land use (CORINE) \+ isolines. 3\) the tree segmentations.*
 
-\institute{}
+## Challenges
 
-\begin{document}
-\maketitle
+Validating large-scale tree crown segmentation models reveals a web of interrelated challenges that go beyond those encountered in small, controlled datasets.
+Seasonal phenology is a moving target: the same forest can look drastically different between leaf-on summer imagery and leaf-off winter scenes.
+Models trained on one season often struggle in another, yielding inconsistent segmentation accuracy across the year.
+For example, a canopy delineation that performs well on lush summer foliage may under-segment sparse autumn crowns or miss bare branches in winter.
+Such season-driven variability not only degrades model performance but also complicates validation – a one-shot model might appear accurate in one season and fail in the next, raising questions about how and when accuracy should be assessed.
+Incorporating multi-season data during both training and validation is essential, as phenological dynamics have been shown to strongly influence model generalizability \[@moussaid2021tree; @cong2022citrus\].
 
-\begin{abstract}
-Advances in deep learning models have opened up new avenues for computer
-vision tasks, like tree crown segmentations from remote sensing images.
-As more and more models become available, the challenges of applying the
-models to really large spatial scales (beyond the forest/plot level)
-become visible. One of these challenges is the validation of the tree
-crown detection and segmentation. Imprecise detections/segmentations can
-lead to inaccuracies in downstream tasks in forestry and urban planning,
-creating a barrier for the application of these models to real-life use
-cases. Attributes of training data, e.g tree sizes, phenological stages,
-number of polygon vertices, shadows, impact the prediction accuracy of
-the segmentation models, hence training data should be chosen with due
-diligence. Here we identify some of the key validation challenges based
-on the collective experiences of not just training deep learning models,
-but also applying the models to very large spatial scales in Germany
-(federal state level).
+Spatial heterogeneity of landscapes, including terrain and illumination differences, poses another major hurdle for both segmentation and its validation.
+An algorithm that segments tree crowns flawlessly in a flat, well-lit park may stumble in a shadow-drenched valley or on a steep mountainside.
+Variations in ground elevation and slope alter the angle of solar illumination, leading to uneven lighting and shadows that can confuse models.
+In mountainous or rugged terrains, trees on north-facing slopes might appear darker or partially occluded compared to those on south-facing slopes with direct sun, even if they are the same species and healthy.
+Such effects result in site-specific performance: models often need fine-tuning or can suffer accuracy drops when moved to new topographies or sensor angles.
+Weinstein et al.
+(2020) observed this kind of cross-site performance gap, where a tree detection model trained in one region underperformed when applied to a different region’s imagery without adaptation, underscoring how terrain and context influence outcomes \[@weinstein2020cross\].
+For validation, this means that accuracy estimates from one area may not transfer to another – a critical issue when assessments at national or global scales are required.
 
-\keywords{First Keyword \and Second Keyword \and Another Keyword.}
+A further fundamental challenge lies in the scarcity of accurate ground-truth data at scale.
+Reliable validation hinges on high-quality reference data (the “ground truth”), yet collecting detailed crown delineations over large regions is logistically difficult and expensive.
+Field surveys can map individual trees with great precision (e.g., measuring trunks and canopy spread on the ground), but doing this over thousands of square kilometers is infeasible.
+Aerial and satellite imagery provide broader coverage for reference data, but even these often lack the resolution or fidelity to unequivocally label each tree crown for validation purposes.
+UAV (drone) campaigns can bridge the gap by capturing very high-resolution images or LiDAR of sample areas, but they are limited in flight range and still require extensive human annotation to turn imagery into usable ground truth.
+The net result is a mismatch of scales: our models aspire to map every tree across entire countries, but our ground truth typically covers only small plots or scattered samples \[^\].
+This mismatch means that validating a “wall-to-wall” tree map often involves extrapolating from a tiny fraction of ground-referenced trees, introducing uncertainty.
+Moreover, ground-reference datasets may not capture the full diversity of conditions (species, canopy shapes, management regimes, etc.) present in the larger mapping area, biasing the validation.
+Expanding ground-truth collection – through automated methods or crowdsourcing – is thus not just a recommendation but a necessity to overcome this validation bottleneck (as we discuss later).
 
-\end{abstract}
-
-\section{Introduction}\label{introduction}
-
-Accurately mapping tree crowns over large areas is critical for scaling
-local observations to regional and global assessments of forest health,
-carbon stocks, and urban greenery. In an era marked by rapid
-environmental change, comprehensive mapping efforts are indispensable.
-Tree crown segmentation is a cornerstone task for these objectives, as
-the size and structure of a tree crown---shaped by species-specific
-branching patterns, site conditions, and competition for
-light---directly influences primary production. A key metric in this
-context is the crown projection area (CPA), defined as the vertical
-projection of the crown onto a horizontal plane. Deriving CPA provides
-essential insights both at the individual tree level---where it provides
-predictions of diameter, volume, and growth rates---and at the stand
-level, supporting models of competition and canopy gap dynamics.
-However, inaccuracies inherent in the segmentation process, particularly
-when applied at large scales, can lead to biased estimations of these
-critical tree variables, underscoring the urgent need to address robust
-validation challenges in large-scale tree crown segmentation. While
-recent advances in deep learning offer unprecedented capabilities for
-automated segmentation, their robust validation at scale presents a
-critical, unaddressed hurdle.
-
-Among the various tree crown variables, crown spread presents unique
-estimation challenges, both in field measurements and, critically, when
-derived from remote sensing imagery. Its accurate derivation is not only
-susceptible to inaccuracies in the initial segmentation process but also
-relies heavily on the methods employed to extract crown spread from
-segmented polygons, a process fraught with its own complexities.
-Selecting an appropriate crown spread calculation method is crucial,
-especially for irregularly shaped tree crown polygons. Even with
-optimized calculation methods, direct matches with ground-truth data
-proved low (e.g., 32\% in our experience), although a broader acceptable
-margin of error (±5 m) captured 89\% of cases. These results immediately
-point to difficulties in establishing precise validation metrics.
-Furthermore, model performance exhibits size-dependent biases: trees
-with small crowns (\textless6 m) are often overestimated, while large
-crowns (\textgreater16 m) tend to be underestimated. These discrepancies
-underscore the inherent limitations of deep learning segmentation models
-at the extremes of the size spectrum. Critically, the quality of
-validation data profoundly influences the assessment of crown spread
-accuracy. For instance, our experience with the local tree inventory
-(`Baumkataster' Halle/Saale), which provides only broad crown spread
-ranges (5 m intervals), severely constrains the precision of validation,
-highlighting the need for more granular and reliable ground-truth
-datasets.
-
-The integration of deep learning into remote sensing has opened new ways
-in how we monitor and analyse ecological systems
-{[}\cite{zheng2024review}, \cite{zhao2023systematic}{]}. Tree crown
-segmentation, a task central to assessing forest health, carbon stocks,
-and urban greenery, has especially benefited from advances in
-convolutional neural networks (CNNs) and transformer-based
-architectures. Approaches leveraging high-resolution imagery have
-demonstrated remarkable capabilities in delineating individual tree
-crowns, as evidenced by works such as \cite{weinstein2019individual},
-\cite{khant2025} and \cite{freudenberg2022individual}, illustrating the
-potential for scalable and automated tree mapping. The urgency of
-climate change and biodiversity loss further amplifies the importance of
-large-scale tree crown segmentation. Forests serve as critical carbon
-sinks and biodiversity reservoirs {[}\cite{pan2024enduring}{]}, while
-urban trees provide essential ecosystem services
-{[}\cite{sharma2024urban}{]}. Large-scale segmentation enables
-consistent, detailed monitoring of these vital resources, informing
-sustainable forest management, urban planning, and policy development.
-The ambition to create national and global inventories, such as those
-envisioned in \cite{tolan2024very}, hinges on the reliability of such
-foundational segmentation data.
-
-However, while these deep learning models demonstrate impressive
-performance in controlled environments or with limited-scale datasets,
-their deployment at expansive spatial extents, such as regional or
-national scales, exposes a distinct spectrum of challenges that remain
-largely unaddressed. Foremost among these is validation---a critical yet
-underexplored hurdle in ensuring the reliability and generalizability of
-model outputs. As recent studies have shown, the performance of tree
-crown segmentation models is highly sensitive to the characteristics of
-the training data, such as tree size, species diversity, seasonal
-variations, and image quality {[}\cite{moussaid2021tree};
-\cite{cong2022citrus}{]}. These factors, when coupled with the
-complexities of real-world landscapes, underscore the need for robust
-and scalable validation frameworks.
-
-This Perspective builds on our extensive experience in applying deep
-learning models to large-scale tree crown segmentation across Germany,
-addressing federal state-level landscapes. By synthesizing insights from
-model training, deployment, and validation, we highlight the nuanced
-challenges associated with assessing segmentation accuracy over
-expansive and heterogeneous terrains. These challenges are not mere
-technicalities---they bear significant implications for downstream
-applications in forestry, conservation, and urban planning. Through this
-work, we aim to not only highlight these critical validation hurdles but
-also to foster a dialogue on the methodological innovations required to
-ensure that deep learning advancements in remote sensing translate into
-actionable insights at the scales demanded by global environmental
-challenges.
-
-\section{Case Study: Tree Crown Segmentation in Saxony and
-Saxony-Anhalt,
-Germany}\label{case-study-tree-crown-segmentation-in-saxony-and-saxony-anhalt-germany}
-
-In this case study, the DeepTrees model {[}\cite{khan2025}{]} was
-applied in a one-shot prediction approach using pretrained model weights
-provided by DeepTrees, applied to high-resolution multispectral digital
-orthophoto imagery (DOP20) covering the German federal states of Saxony
-(SN) and Saxony-Anhalt (ST). Sachsen, covering approximately 18,450 km²,
-and Saxony-Anhalt, spanning around 20,452 km², represent diverse
-ecological and urban landscapes, ideal for assessing large-scale
-segmentation model performance.
-
-DOP20 imagery provides detailed spatial information at 20 cm resolution
-per pixel, capturing visual and near infrared spectral regions essential
-for accurate tree crown delineation. Using these pretrained weights, the
-DeepTrees model identified approximately 218.7 million individual tree
-crowns---137.3 million in Saxony and 81.4 million in Saxony-Anhalt. The
-resulting segmentation dataset has been made publicly available on
-Zenodo (https://zenodo.org/record/exampledoi), facilitating further
-research and validation.
-
-This comprehensive segmentation highlights substantial regional
-variability in tree distribution, reflecting ecological, topographical,
-and land-use gradients. The case study thus forms a practical foundation
-for addressing the validation challenges detailed subsequently,
-emphasizing the necessity for robust validation frameworks to ensure
-reliable ecological monitoring and informed decision-making based on
-large-scale automated analyses.
-
-\emph{Figure 1: 6 panels (1 for each state). From left to right
--\textgreater{} 1) tiles with date classification 2) Land use (CORINE) +
-isolines. 3) the tree segmentations.}
-
-\section{Challenges}\label{challenges}
-
-Validating large-scale tree crown segmentation models reveals a web of
-interrelated challenges that go beyond those encountered in small,
-controlled datasets. Seasonal phenology is a moving target: the same
-forest can look drastically different between leaf-on summer imagery and
-leaf-off winter scenes. Models trained on one season often struggle in
-another, yielding inconsistent segmentation accuracy across the year.
-For example, a canopy delineation that performs well on lush summer
-foliage may under-segment sparse autumn crowns or miss bare branches in
-winter. Such season-driven variability not only degrades model
-performance but also complicates validation -- a one-shot model might
-appear accurate in one season and fail in the next, raising questions
-about how and when accuracy should be assessed. Incorporating
-multi-season data during both training and validation is essential, as
-phenological dynamics have been shown to strongly influence model
-generalizability {[}\cite{moussaid2021tree}; \cite{cong2022citrus}{]}.
-
-Spatial heterogeneity of landscapes, including terrain and illumination
-differences, poses another major hurdle for both segmentation and its
-validation. An algorithm that segments tree crowns flawlessly in a flat,
-well-lit park may stumble in a shadow-drenched valley or on a steep
-mountainside. Variations in ground elevation and slope alter the angle
-of solar illumination, leading to uneven lighting and shadows that can
-confuse models. In mountainous or rugged terrains, trees on north-facing
-slopes might appear darker or partially occluded compared to those on
-south-facing slopes with direct sun, even if they are the same species
-and healthy. Such effects result in site-specific performance: models
-often need fine-tuning or can suffer accuracy drops when moved to new
-topographies or sensor angles. Weinstein et al.~(2020) observed this
-kind of cross-site performance gap, where a tree detection model trained
-in one region underperformed when applied to a different region's
-imagery without adaptation, underscoring how terrain and context
-influence outcomes {[}\cite{weinstein2020cross}{]}. For validation, this
-means that accuracy estimates from one area may not transfer to another
--- a critical issue when assessments at national or global scales are
-required.
-
-A further fundamental challenge lies in the scarcity of accurate
-ground-truth data at scale. Reliable validation hinges on high-quality
-reference data (the ``ground truth''), yet collecting detailed crown
-delineations over large regions is logistically difficult and expensive.
-Field surveys can map individual trees with great precision (e.g.,
-measuring trunks and canopy spread on the ground), but doing this over
-thousands of square kilometers is infeasible. Aerial and satellite
-imagery provide broader coverage for reference data, but even these
-often lack the resolution or fidelity to unequivocally label each tree
-crown for validation purposes. UAV (drone) campaigns can bridge the gap
-by capturing very high-resolution images or LiDAR of sample areas, but
-they are limited in flight range and still require extensive human
-annotation to turn imagery into usable ground truth. The net result is a
-mismatch of scales: our models aspire to map every tree across entire
-countries, but our ground truth typically covers only small plots or
-scattered samples {[}\^{}{]}. This mismatch means that validating a
-``wall-to-wall'' tree map often involves extrapolating from a tiny
-fraction of ground-referenced trees, introducing uncertainty. Moreover,
-ground-reference datasets may not capture the full diversity of
-conditions (species, canopy shapes, management regimes, etc.) present in
-the larger mapping area, biasing the validation. Expanding ground-truth
-collection -- through automated methods or crowdsourcing -- is thus not
-just a recommendation but a necessity to overcome this validation
-bottleneck (as we discuss later).
-
-Compounding the issue of limited data is the inconsistency in reference
-annotations and evaluation metrics. Even when ground-reference data
-exist, their format can differ -- sometimes reference trees are marked
-by a single GPS point (e.g., trunk location), sometimes by a hand-drawn
-polygon outlining the crown. This creates a challenge in validation: how
-do we decide if a predicted crown polygon ``matches'' a ground-truth
-point, or how to handle cases where one field-mapped tree corresponds to
-multiple overlapping crown segments in the image?(Figure ref)
-Conversely, field crews might delineate a broad canopy as one crown
-while an automated model splits it into two segments (or vice versa),
-especially in dense stands where crowns merge. These ambiguities in
-one-to-one correspondence make it hard to define what a ``correct''
-segmentation is. Traditional pixel-wise accuracy metrics like
-Intersection-over-Union (IoU) treat segmentation purely as an image
-overlap problem, which may not reflect the ecological reality of
-counting individual trees. IoU penalizes differences in shape or area
-but doesn't account for whether the count of tree objects is correct. In
-an extreme case, a model could slightly over-segment every tree
-(splitting each true crown into two smaller polygons) and still achieve
-a reasonable IoU, despite doubling the perceived tree count -- a
-significant error for applications. On the other hand, object-centric
-metrics such as panoptic segmentation quality attempt to consider both
-detection and delineation of objects
-{[}\cite{kirillov2019panopticsegmentation}{]}. Panoptic metrics combine
-aspects of object detection (was each tree detected?) with segmentation
-quality (was each crown correctly outlined?), which can be more
-appropriate for tree mapping. However, even these require well-defined
-ground-truth objects to compare against. When the ground truth itself is
-inconsistent (e.g., how to count a clumped cluster of stems with
-overlapping crowns), validation metrics struggle to fully capture model
-performance. The choice of evaluation metric thus becomes non-trivial:
-depending on whether one prioritizes exact crown shape, tree count, or
-canopy cover, the ``best'' metric may differ. Establishing consensus on
-evaluation protocols is part of the challenge -- without it, different
-studies may report accuracy in incompatible ways.
+Compounding the issue of limited data is the inconsistency in reference annotations and evaluation metrics.
+Even when ground-reference data exist, their format can differ – sometimes reference trees are marked by a single GPS point (e.g., trunk location), sometimes by a hand-drawn polygon outlining the crown.
+This creates a challenge in validation: how do we decide if a predicted crown polygon “matches” a ground-truth point, or how to handle cases where one field-mapped tree corresponds to multiple overlapping crown segments in the image?(Figure ref)  Conversely, field crews might delineate a broad canopy as one crown while an automated model splits it into two segments (or vice versa), especially in dense stands where crowns merge.
+These ambiguities in one-to-one correspondence make it hard to define what a “correct” segmentation is.
+Traditional pixel-wise accuracy metrics like Intersection-over-Union (IoU) treat segmentation purely as an image overlap problem, which may not reflect the ecological reality of counting individual trees.
+IoU penalizes differences in shape or area but doesn’t account for whether the count of tree objects is correct.
+In an extreme case, a model could slightly over-segment every tree (splitting each true crown into two smaller polygons) and still achieve a reasonable IoU, despite doubling the perceived tree count – a significant error for applications.
+On the other hand, object-centric metrics such as panoptic segmentation quality attempt to consider both detection and delineation of objects \[@kirillov2019panopticsegmentation\].
+Panoptic metrics combine aspects of object detection (was each tree detected?) with segmentation quality (was each crown correctly outlined?), which can be more appropriate for tree mapping.
+However, even these require well-defined ground-truth objects to compare against.
+When the ground truth itself is inconsistent (e.g., how to count a clumped cluster of stems with overlapping crowns), validation metrics struggle to fully capture model performance.
+The choice of evaluation metric thus becomes non-trivial: depending on whether one prioritizes exact crown shape, tree count, or canopy cover, the “best” metric may differ.
+Establishing consensus on evaluation protocols is part of the challenge – without it, different studies may report accuracy in incompatible ways.
 
 There is also the issue of scale and resolution in validation reporting.
-A model's accuracy can appear to vary depending on the spatial scale at
-which it's evaluated. For instance, a segmentation model might achieve
-high overall accuracy when averaged over an entire region, yet if one
-zooms into a small test area (say a single city park or forest stand),
-the error rate might be much higher or lower. This can happen if errors
-are not evenly distributed: the model could perform very well in one
-type of landscape (e.g., neat urban street trees) and poorly in another
-(dense natural forest), and a coarse regional average could mask these
-extremes. Consequently, a user working on a local conservation project
-might experience worse performance than the ``headline'' accuracy
-suggests, because that headline number was diluted by many easier cases
-elsewhere. Ensuring that validation is robust across scales is tricky --
-one must balance broad coverage with local detail. It calls for
-multi-scale validation approaches, where accuracy is reported at
-multiple grain sizes or stratified by landscape type. Highlighting this,
-one could imagine a figure plotting model accuracy as a function of
-spatial extent or across different habitat types, illustrating how
-performance can drop off in specific challenging subsets despite looking
-good overall. This emphasizes that heterogeneous performance is itself a
-challenge to acknowledge in validation: we need methods to detect where
-and why a model fails, not just an aggregate score.
+A model’s accuracy can appear to vary depending on the spatial scale at which it’s evaluated.
+For instance, a segmentation model might achieve high overall accuracy when averaged over an entire region, yet if one zooms into a small test area (say a single city park or forest stand), the error rate might be much higher or lower.
+This can happen if errors are not evenly distributed: the model could perform very well in one type of landscape (e.g., neat urban street trees) and poorly in another (dense natural forest), and a coarse regional average could mask these extremes.
+Consequently, a user working on a local conservation project might experience worse performance than the “headline” accuracy suggests, because that headline number was diluted by many easier cases elsewhere.
+Ensuring that validation is robust across scales is tricky – one must balance broad coverage with local detail.
+It calls for multi-scale validation approaches, where accuracy is reported at multiple grain sizes or stratified by landscape type.
+Highlighting this, one could imagine a figure plotting model accuracy as a function of spatial extent or across different habitat types, illustrating how performance can drop off in specific challenging subsets despite looking good overall.
+This emphasizes that heterogeneous performance is itself a challenge to acknowledge in validation: we need methods to detect where and why a model fails, not just an aggregate score.
 
-To summarize, validating large-scale tree segmentation models is far
-from a straightforward task. Seasonal changes, diverse terrain and
-illumination, limited and inconsistent ground truth, ambiguous
-evaluation criteria, and scale-dependent performance all intertwine to
-create a demanding setting. These challenges are not just academic --
-they directly impact how much trust we can place in AI-generated tree
-maps for real-world decisions in forestry, ecology, and urban planning.
-Recognizing these pain points is the first step; the next is devising
-strategies to overcome them. We now turn to several recommendations
-aimed at improving both models and validation practices, with an eye
-toward bridging the gap between controlled experiments and the
-complexity of continental-scale deployments.
+To summarize, validating large-scale tree segmentation models is far from a straightforward task.
+Seasonal changes, diverse terrain and illumination, limited and inconsistent ground truth, ambiguous evaluation criteria, and scale-dependent performance all intertwine to create a demanding setting.
+These challenges are not just academic – they directly impact how much trust we can place in AI-generated tree maps for real-world decisions in forestry, ecology, and urban planning.
+Recognizing these pain points is the first step; the next is devising strategies to overcome them.
+We now turn to several recommendations aimed at improving both models and validation practices, with an eye toward bridging the gap between controlled experiments and the complexity of continental-scale deployments.
 
-\emph{Table 1: Potential validation challenges and corresponding
-mitigation strategies. This table could list key challenges in
-validating large-scale tree segmentations alongside proposed approaches
-to address them. For example: (Challenge) Phenological variation between
-seasons -- (Mitigation) use multi-temporal training data and
-season-specific validation sets; (Challenge) Heterogeneous terrain and
-shadowing -- (Mitigation) integrate digital elevation models and
-terrain-aware modeling; (Challenge) Scarcity of ground-truth annotations
--- (Mitigation) leverage semi-automated labeling, crowdsourcing, and
-active learning to expand validation data; (Challenge) Ambiguity in
-crown delineation -- (Mitigation) adopt object-based accuracy metrics
-and consensus protocols for what counts as a single tree; (Challenge)
-Scale-dependent performance -- (Mitigation) evaluate models at multiple
-spatial scales and stratify results by landscape type. This overview
-would visually reinforce the narrative that each challenge has an
-identifiable path forward, linking the ``Challenges'' and
-``Recommendations'' sections.}
+*Table 1: Potential validation challenges and corresponding mitigation strategies.
+This table could list key challenges in validating large-scale tree segmentations alongside proposed approaches to address them.
+For example: (Challenge) Phenological variation between seasons – (Mitigation) use multi-temporal training data and season-specific validation sets; (Challenge) Heterogeneous terrain and shadowing – (Mitigation) integrate digital elevation models and terrain-aware modeling; (Challenge) Scarcity of ground-truth annotations – (Mitigation) leverage semi-automated labeling, crowdsourcing, and active learning to expand validation data; (Challenge) Ambiguity in crown delineation – (Mitigation) adopt object-based accuracy metrics and consensus protocols for what counts as a single tree; (Challenge) Scale-dependent performance – (Mitigation) evaluate models at multiple spatial scales and stratify results by landscape type.
+This overview would visually reinforce the narrative that each challenge has an identifiable path forward, linking the “Challenges” and “Recommendations” sections.*
 
-\emph{!{[}{]}{[}image1{]}}\\
-\emph{Figure 2: Incomplete}
+*![][image1]*  
+*Figure 2: Incomplete*
 
-\section{Recommendations}\label{recommendations}
+## Recommendations
 
-Addressing the above challenges requires a multi-pronged approach,
-combining improvements in modeling techniques with innovations in
-validation methodology and data collection. We outline here several
-complementary strategies that, together, can bolster the reliability of
-large-scale tree segmentation efforts. These recommendations emphasize
-building more robust models through advanced training paradigms, as well
-as creating better frameworks to evaluate and support those models in
-real-world conditions. The overarching goal is to ensure that deep
-learning advancements in tree mapping translate into trustworthy,
-actionable insights at regional to global scales.
+Addressing the above challenges requires a multi-pronged approach, combining improvements in modeling techniques with innovations in validation methodology and data collection.
+We outline here several complementary strategies that, together, can bolster the reliability of large-scale tree segmentation efforts.
+These recommendations emphasize building more robust models through advanced training paradigms, as well as creating better frameworks to evaluate and support those models in real-world conditions.
+The overarching goal is to ensure that deep learning advancements in tree mapping translate into trustworthy, actionable insights at regional to global scales.
 
-\textbf{Leverage self-supervised learning for robust representations.} A
-key step towards more generalizable segmentation models is tapping into
-unannotated data via self-supervised learning (SSL). Unlike traditional
-supervised training which is bottlenecked by limited labeled examples,
-SSL allows models to learn from the abundant pool of unlabeled remote
-sensing images -- for example, by predicting missing pieces of an image
-or distinguishing augmented views of the same scene. By pre-training a
-model on large geospatial datasets without any human-provided labels,
-the model can absorb intrinsic patterns of the landscape: textures,
-shapes, seasonal changes, and other context that are common across
-images. These rich foundational representations can then be fine-tuned
-for tree crown segmentation with far fewer annotated samples than would
-otherwise be needed. Recent efforts like PhilEO Bench demonstrate the
-promise of this approach, evaluating geospatial foundation models that
-were pre-trained using SSL techniques {[}\cite{fibaek2024phileo}{]}. The
-results show improved performance on a range of tasks (e.g.~building
-footprint extraction, road mapping) after such pre-training, compared to
-models trained from scratch. In the context of trees, a model with an
-SSL-pretrained backbone may already ``know'' about basic vegetation
-structures, shadows, and seasonal appearances, making it more adept at
-delineating crowns under varied conditions. For instance, a foundation
-model trained on year-round satellite imagery might implicitly
-understand the difference between a leafless oak in winter and the same
-oak in summer, and thus require only a light fine-tuning to accurately
-segment each. Mendieta et al.~(2023) take this further by continually
-pre-training on new data distributions (a form of continual learning),
-which helped build a Geospatial Foundation Model (GFM) that excelled
-across multiple remote sensing tasks {[}\cite{mendieta2023gfm}{]}. Such
-continual SSL training could allow segmentation models to keep improving
-as more unlabeled data (e.g., new satellite images over time or from new
-regions) become available, staying up-to-date with changes in landscapes
-and sensor characteristics. Exploiting SSL is a powerful way to tackle
-the twin issues of limited labels and dataset bias, yielding a model
-that is better equipped for the diverse scenarios encountered in
-large-scale mapping.
+**Leverage self-supervised learning for robust representations.** A key step towards more generalizable segmentation models is tapping into unannotated data via self-supervised learning (SSL).
+Unlike traditional supervised training which is bottlenecked by limited labeled examples, SSL allows models to learn from the abundant pool of unlabeled remote sensing images – for example, by predicting missing pieces of an image or distinguishing augmented views of the same scene.
+By pre-training a model on large geospatial datasets without any human-provided labels, the model can absorb intrinsic patterns of the landscape: textures, shapes, seasonal changes, and other context that are common across images.
+These rich foundational representations can then be fine-tuned for tree crown segmentation with far fewer annotated samples than would otherwise be needed.
+Recent efforts like PhilEO Bench demonstrate the promise of this approach, evaluating geospatial foundation models that were pre-trained using SSL techniques \[@fibaek2024phileo\].
+The results show improved performance on a range of tasks (e.g.
+building footprint extraction, road mapping) after such pre-training, compared to models trained from scratch.
+In the context of trees, a model with an SSL-pretrained backbone may already “know” about basic vegetation structures, shadows, and seasonal appearances, making it more adept at delineating crowns under varied conditions.
+For instance, a foundation model trained on year-round satellite imagery might implicitly understand the difference between a leafless oak in winter and the same oak in summer, and thus require only a light fine-tuning to accurately segment each.
+Mendieta et al.
+(2023) take this further by continually pre-training on new data distributions (a form of continual learning), which helped build a Geospatial Foundation Model (GFM) that excelled across multiple remote sensing tasks \[@mendieta2023gfm\].
+Such continual SSL training could allow segmentation models to keep improving as more unlabeled data (e.g., new satellite images over time or from new regions) become available, staying up-to-date with changes in landscapes and sensor characteristics.
+Exploiting SSL is a powerful way to tackle the twin issues of limited labels and dataset bias, yielding a model that is better equipped for the diverse scenarios encountered in large-scale mapping.
 
-\textbf{Integrate multi-view and multi-temporal data for consistency.}
-Another promising avenue is to train and evaluate models with multi-view
-inputs -- that is, imagery of the same trees captured from different
-angles, sensors, or times. Multi-view here can mean different things:
-multi-angle (such as combining nadir and oblique aerial images),
-multi-platform (satellite imagery plus drone photos), or multi-temporal
-(images taken in different seasons or years). By exposing models to such
-spatiotemporal diversity, we can help them learn invariances that make
-segmentation more reliable. For example, a tree crown seen from directly
-above might have a certain shape, but from a side angle the outline
-might blend with neighbors; a model trained with both perspectives could
-learn to robustly identify the crown in either view. Likewise, combining
-leaf-on and leaf-off images of the same forest during training can force
-a model to rely on structural features (branches, trunk hints, relative
-spacing) in addition to just greenness, thereby improving its ability to
-generalize across seasons. Self-supervised methods are particularly
-well-suited to capitalize on multi-view data because they can be set up
-to encourage the model to produce similar latent representations for
-different views of the same object. Techniques such as masked
-autoencoders or contrastive learning can be used on multi-view datasets
-to make a model predict or align one view from another, without any
-manual labels needed. Such approaches have already shown success: for
-instance, researchers have used masked image modeling and contrastive
-SSL on multi-view satellite imagery to boost performance on segmentation
-and detection tasks, essentially teaching the model that ``these two
-different-looking images actually contain the same trees''
-{[}\cite{mukkavilli2023prithvi}; \cite{fibaek2024phileo}{]}. By training
-with multi-view consistency, the model becomes more robust to viewpoint
-and temporal changes, which directly addresses the phenology and
-illumination challenges discussed earlier. Importantly, this not only
-helps the model's predictions but also strengthens validation, because a
-model that is consistent across views makes it easier to compare
-predicted and true crowns even when the reference data comes from a
-slightly different perspective or date. In practice, one could imagine a
-validation scheme where the agreement of a model's output between
-leaf-on and leaf-off images serves as an indicator of reliability: large
-discrepancies might flag areas for further human inspection. Overall,
-multi-view and multi-temporal training imbue models with a form of
-contextual intelligence about the 3D and time-varying nature of trees,
-making segmentation outcomes more stable across the real-world
-variability that large-scale applications inevitably encompass.
+**Integrate multi-view and multi-temporal data for consistency.** Another promising avenue is to train and evaluate models with multi-view inputs – that is, imagery of the same trees captured from different angles, sensors, or times.
+Multi-view here can mean different things: multi-angle (such as combining nadir and oblique aerial images), multi-platform (satellite imagery plus drone photos), or multi-temporal (images taken in different seasons or years).
+By exposing models to such spatiotemporal diversity, we can help them learn invariances that make segmentation more reliable.
+For example, a tree crown seen from directly above might have a certain shape, but from a side angle the outline might blend with neighbors; a model trained with both perspectives could learn to robustly identify the crown in either view.
+Likewise, combining leaf-on and leaf-off images of the same forest during training can force a model to rely on structural features (branches, trunk hints, relative spacing) in addition to just greenness, thereby improving its ability to generalize across seasons.
+Self-supervised methods are particularly well-suited to capitalize on multi-view data because they can be set up to encourage the model to produce similar latent representations for different views of the same object.
+Techniques such as masked autoencoders or contrastive learning can be used on multi-view datasets to make a model predict or align one view from another, without any manual labels needed.
+Such approaches have already shown success: for instance, researchers have used masked image modeling and contrastive SSL on multi-view satellite imagery to boost performance on segmentation and detection tasks, essentially teaching the model that “these two different-looking images actually contain the same trees” \[@mukkavilli2023prithvi; @fibaek2024phileo\].
+By training with multi-view consistency, the model becomes more robust to viewpoint and temporal changes, which directly addresses the phenology and illumination challenges discussed earlier.
+Importantly, this not only helps the model’s predictions but also strengthens validation, because a model that is consistent across views makes it easier to compare predicted and true crowns even when the reference data comes from a slightly different perspective or date.
+In practice, one could imagine a validation scheme where the agreement of a model’s output between leaf-on and leaf-off images serves as an indicator of reliability: large discrepancies might flag areas for further human inspection.
+Overall, multi-view and multi-temporal training imbue models with a form of contextual intelligence about the 3D and time-varying nature of trees, making segmentation outcomes more stable across the real-world variability that large-scale applications inevitably encompass.
 
-\textbf{Incorporate terrain information into the modeling pipeline}. As
-noted, uneven terrain can cause substantial variability in how trees
-appear in images, so bringing explicit knowledge of terrain into the
-segmentation process is a logical remedy. One recommendation is to fuse
-digital elevation models (DEMs) or LiDAR-derived terrain data with the
-imagery during model training. This could be as simple as providing the
-model with an extra channel of input encoding elevation/slope, or as
-complex as designing the model to separately process terrain context.
-Self-supervised pre-training can be extended here too: recent work on
-multisensor geospatial foundation models has shown that including
-elevation data in SSL (e.g., tasking the model to distinguish imagery of
-flat vs.~mountainous areas) yields better feature representations for
-downstream tasks {[}\cite{han2024msGFM}{]}. By differentiating between
-bare earth and above-ground structures in pre-training, the model learns
-to discount illumination differences that are purely due to slope and
-aspect, focusing instead on actual objects like trees. In a tree
-segmentation scenario, a terrain-informed model could recognize that a
-dark region in an image is a shaded hillside rather than a non-existent
-gap in canopy cover, or that an elongated shape on a steep slope is
-still a single tree crown albeit skewed by perspective. Incorporating
-terrain data directly addresses the challenge of spatial heterogeneity:
-it provides a reference frame to normalize out some variability. This
-can improve validation as well -- for example, error analysis can be
-stratified by terrain class to ensure a model works not just on average,
-but on hilltops and valleys alike. We recommend that future segmentation
-models, especially for large regions with varied topography, adopt
-terrain-aware training strategies. Even if a full DEM is not available
-everywhere, approximating slope from the imagery or using coarse global
-elevation data could still offer benefits. Ultimately, bridging the gap
-between pixel appearance and real-world topography makes the model's
-understanding more physical and generalizable, reducing surprises when
-it's deployed on a new landscape.
+**Incorporate terrain information into the modeling pipeline**.
+As noted, uneven terrain can cause substantial variability in how trees appear in images, so bringing explicit knowledge of terrain into the segmentation process is a logical remedy.
+One recommendation is to fuse digital elevation models (DEMs) or LiDAR-derived terrain data with the imagery during model training.
+This could be as simple as providing the model with an extra channel of input encoding elevation/slope, or as complex as designing the model to separately process terrain context.
+Self-supervised pre-training can be extended here too: recent work on multisensor geospatial foundation models has shown that including elevation data in SSL (e.g., tasking the model to distinguish imagery of flat vs.
+mountainous areas) yields better feature representations for downstream tasks \[@han2024msGFM\].
+By differentiating between bare earth and above-ground structures in pre-training, the model learns to discount illumination differences that are purely due to slope and aspect, focusing instead on actual objects like trees.
+In a tree segmentation scenario, a terrain-informed model could recognize that a dark region in an image is a shaded hillside rather than a non-existent gap in canopy cover, or that an elongated shape on a steep slope is still a single tree crown albeit skewed by perspective.
+Incorporating terrain data directly addresses the challenge of spatial heterogeneity: it provides a reference frame to normalize out some variability.
+This can improve validation as well – for example, error analysis can be stratified by terrain class to ensure a model works not just on average, but on hilltops and valleys alike.
+We recommend that future segmentation models, especially for large regions with varied topography, adopt terrain-aware training strategies.
+Even if a full DEM is not available everywhere, approximating slope from the imagery or using coarse global elevation data could still offer benefits.
+Ultimately, bridging the gap between pixel appearance and real-world topography makes the model’s understanding more physical and generalizable, reducing surprises when it’s deployed on a new landscape.
 
-\textbf{Focus model attention on domain-specific features.} Beyond data
-augmentation and multi-source inputs, improvements in the model
-architecture and training objectives themselves can yield more reliable
-segmentation. One intriguing direction is the use of feature-guided
-masked autoencoders or similar techniques that encourage the model to
-learn high-level features rather than getting bogged down in pixel-level
-noise. In remote sensing, not all pixels are equal -- the spectral
-signature of a healthy tree canopy, for instance, is characterized by
-certain reflectance patterns (like high near-infrared reflectance for
-foliage), and textural cues can differentiate a tree from grass. A
-vanilla model might or might not latch onto these subtle cues. However,
-a feature-guided approach explicitly trains the model to reconstruct or
-predict meaningful feature representations (such as vegetation indices
-or edge maps) instead of raw pixel intensity. For example, a recent
-approach called FG-MAE (Feature Guided Masked Autoencoder) masks out
-parts of an image and tasks the model with predicting domain-relevant
-features (like a NDVI -- Normalized Difference Vegetation Index image,
-or other engineered representations) for the masked region rather than
-the raw pixels. This forces the model to infer what type of object
-should be there, not just to copy textures, thereby learning a more
-semantic understanding of the scene {[}\cite{mukkavilli2023prithvi}{]}.
-Applying this idea to tree segmentation, we could pre-train models to
-predict features that highlight vegetation structure (perhaps oriented
-gradients that capture tree crown edges, or canopy height estimations
-from stereo images) so that the model's internal representations become
-highly attuned to ``tree-ness.'' When fine-tuned for segmentation, such
-a model may be better at delineating tangled canopies or differentiating
-trees from confusing background elements, because it has learned to
-focus on the attributes that define a tree in imagery. Early studies in
-multispectral and SAR domains have found that this approach yields
-improved performance in complex environments
-{[}\cite{mukkavilli2023prithvi}{]}. We recommend integrating
-feature-guidance in training for large-scale tree mapping, especially in
-areas with complex backgrounds (e.g., urban environments where trees
-mingle with buildings). Not only does this likely boost accuracy, it
-could also produce more interpretable model outputs or uncertainties --
-a model that knows what features it's looking for might provide
-reasoning (explicit or implicit) for its segmentation, which in turn
-aids validation and error diagnosis.
+**Focus model attention on domain-specific features.** Beyond data augmentation and multi-source inputs, improvements in the model architecture and training objectives themselves can yield more reliable segmentation.
+One intriguing direction is the use of feature-guided masked autoencoders or similar techniques that encourage the model to learn high-level features rather than getting bogged down in pixel-level noise.
+In remote sensing, not all pixels are equal – the spectral signature of a healthy tree canopy, for instance, is characterized by certain reflectance patterns (like high near-infrared reflectance for foliage), and textural cues can differentiate a tree from grass.
+A vanilla model might or might not latch onto these subtle cues.
+However, a feature-guided approach explicitly trains the model to reconstruct or predict meaningful feature representations (such as vegetation indices or edge maps) instead of raw pixel intensity.
+For example, a recent approach called FG-MAE (Feature Guided Masked Autoencoder) masks out parts of an image and tasks the model with predicting domain-relevant features (like a NDVI – Normalized Difference Vegetation Index image, or other engineered representations) for the masked region rather than the raw pixels.
+This forces the model to infer what type of object should be there, not just to copy textures, thereby learning a more semantic understanding of the scene \[@mukkavilli2023prithvi\].
+Applying this idea to tree segmentation, we could pre-train models to predict features that highlight vegetation structure (perhaps oriented gradients that capture tree crown edges, or canopy height estimations from stereo images) so that the model’s internal representations become highly attuned to “tree-ness.” When fine-tuned for segmentation, such a model may be better at delineating tangled canopies or differentiating trees from confusing background elements, because it has learned to focus on the attributes that define a tree in imagery.
+Early studies in multispectral and SAR domains have found that this approach yields improved performance in complex environments \[@mukkavilli2023prithvi\].
+We recommend integrating feature-guidance in training for large-scale tree mapping, especially in areas with complex backgrounds (e.g., urban environments where trees mingle with buildings).
+Not only does this likely boost accuracy, it could also produce more interpretable model outputs or uncertainties – a model that knows what features it’s looking for might provide reasoning (explicit or implicit) for its segmentation, which in turn aids validation and error diagnosis.
 
-\textbf{Expand validation beyond pixel agreement -- use ecological
-consistency checks.} Traditional validation of segmentation focuses on
-geometric overlap with ground truth shapes, but for tree mapping we can
-also exploit well-established ecological relationships as an additional
-form of validation. Trees have allometric relationships -- mathematical
-links between dimensions like trunk diameter, height, crown diameter,
-and biomass. These relationships have been measured in forestry for
-decades. For example, a tree of a given height typically has a crown of
-roughly proportional diameter, and there are known bounds on how big a
-crown can get for a given trunk size. We propose using such allometric
-equations as a sanity check for segmentation outputs. If a model's
-predicted tree crowns violate basic allometry (say, a tiny tree height
-but a huge crown width, or a cluster of crowns whose total basal area
-implies an impossible biomass for that area), it might indicate errors
-in the segmentation or missing trees. One could quantitatively compare
-the distribution of predicted crown sizes and tree heights (if height
-data or estimates are available) against expected distributions from
-field data. Significant deviations could flag problems: for instance, if
-the model frequently delineates very large crowns that in reality would
-correspond to 80-meter tall trees (which don't exist in the region),
-those are likely over-segmentation artifacts. Conversely, if predicted
-crowns are all very small in an old-growth forest where trees should be
-large, the model might be under-segmenting (splitting one crown into
-many). Researchers have indeed used this kind of approach in related
-contexts; for example, Song et al.~(2023) employed Gaussian process
-regression on tree height to estimate biomass, demonstrating how linking
-remote sensing outputs to allometric models can validate whether the
-outputs make ecological sense {[}\cite{song2023biomass}{]}. In practice,
-implementing allometric checks means bringing in additional data or
-models (e.g., a LiDAR-derived tree height map, or species-specific
-allometric formulas from forestry literature) and cross-verifying the
-plausibility of the AI-generated tree map. This recommendation shifts
-validation from a purely computer-vision perspective to an
-application-oriented perspective: after all, if the ultimate goal is to
-use these maps for carbon accounting, biodiversity, or forestry, then
-passing an ecological reality check is as important as scoring well on
-IoU.
+**Expand validation beyond pixel agreement – use ecological consistency checks.** Traditional validation of segmentation focuses on geometric overlap with ground truth shapes, but for tree mapping we can also exploit well-established ecological relationships as an additional form of validation.
+Trees have allometric relationships – mathematical links between dimensions like trunk diameter, height, crown diameter, and biomass.
+These relationships have been measured in forestry for decades.
+For example, a tree of a given height typically has a crown of roughly proportional diameter, and there are known bounds on how big a crown can get for a given trunk size.
+We propose using such allometric equations as a sanity check for segmentation outputs.
+If a model’s predicted tree crowns violate basic allometry (say, a tiny tree height but a huge crown width, or a cluster of crowns whose total basal area implies an impossible biomass for that area), it might indicate errors in the segmentation or missing trees.
+One could quantitatively compare the distribution of predicted crown sizes and tree heights (if height data or estimates are available) against expected distributions from field data.
+Significant deviations could flag problems: for instance, if the model frequently delineates very large crowns that in reality would correspond to 80-meter tall trees (which don’t exist in the region), those are likely over-segmentation artifacts.
+Conversely, if predicted crowns are all very small in an old-growth forest where trees should be large, the model might be under-segmenting (splitting one crown into many).
+Researchers have indeed used this kind of approach in related contexts; for example, Song et al.
+(2023) employed Gaussian process regression on tree height to estimate biomass, demonstrating how linking remote sensing outputs to allometric models can validate whether the outputs make ecological sense \[@song2023biomass\].
+In practice, implementing allometric checks means bringing in additional data or models (e.g., a LiDAR-derived tree height map, or species-specific allometric formulas from forestry literature) and cross-verifying the plausibility of the AI-generated tree map.
+This recommendation shifts validation from a purely computer-vision perspective to an application-oriented perspective: after all, if the ultimate goal is to use these maps for carbon accounting, biodiversity, or forestry, then passing an ecological reality check is as important as scoring well on IoU.
 
-\textbf{Establish community benchmarks and evaluation frameworks.} The
-field of geospatial AI is recognizing the value of standard benchmarks
--- datasets and metrics on which different methods can be compared in a
-reproducible way. For computer vision in general, ImageNet and COCO
-served this role, and now geospatial analogues are emerging. Efforts
-like GEO-Bench and PANGAEA have started assembling diverse geospatial
-tasks into comprehensive evaluation suites, underscoring the importance
-of benchmarking models across varying conditions
-{[}\cite{marsocci2024pangaea}{]}. We recommend developing similar
-benchmark datasets and challenges specifically for tree crown
-segmentation and mapping. These benchmarks should cover a wide range of
-landscapes (different forest types, urban trees, plantations, etc.),
-seasons, and remote sensing data sources, reflecting the real-world
-diversity highlighted in our Challenges section. Importantly, a common
-benchmark would promote consensus on validation metrics -- if the
-community agrees to evaluate on, say, a mix of IoU for segmentation
-quality and perhaps a detection metric for counting accuracy, it would
-standardize how results are reported. This comparability accelerates
-progress: researchers can pinpoint which innovations truly lead to
-better performance and robustness. Moreover, a shared evaluation
-framework can include protocols for multi-scale assessment (e.g.,
-requiring methods to report not just overall accuracy but also
-performance on hard subsets like ``dense forest canopy'' vs ``isolated
-trees'') and encourage inclusion of auxiliary checks like the allometric
-consistency mentioned above. In short, a benchmarking initiative for
-large-scale tree segmentation would provide an invaluable feedback loop,
-where challenges discovered by one team (e.g., failure cases in a
-certain region or condition) become part of the test set that everyone
-then tries to solve. This kind of iterative improvement cycle has been
-crucial in other domains of AI. We envision something like an annual
-challenge where models compete on a standard large-scale tree mapping
-task -- pushing methods to be not only accurate but also general and
-validation-friendly. By establishing these community standards, we move
-toward a future where claims of model performance are transparent and
-believable, because they've been vetted on a broad, agreed-upon spectrum
-of scenarios.
+**Establish community benchmarks and evaluation frameworks.** The field of geospatial AI is recognizing the value of standard benchmarks – datasets and metrics on which different methods can be compared in a reproducible way.
+For computer vision in general, ImageNet and COCO served this role, and now geospatial analogues are emerging.
+Efforts like GEO-Bench and PANGAEA have started assembling diverse geospatial tasks into comprehensive evaluation suites, underscoring the importance of benchmarking models across varying conditions \[@marsocci2024pangaea\].
+We recommend developing similar benchmark datasets and challenges specifically for tree crown segmentation and mapping.
+These benchmarks should cover a wide range of landscapes (different forest types, urban trees, plantations, etc.), seasons, and remote sensing data sources, reflecting the real-world diversity highlighted in our Challenges section.
+Importantly, a common benchmark would promote consensus on validation metrics – if the community agrees to evaluate on, say, a mix of IoU for segmentation quality and perhaps a detection metric for counting accuracy, it would standardize how results are reported.
+This comparability accelerates progress: researchers can pinpoint which innovations truly lead to better performance and robustness.
+Moreover, a shared evaluation framework can include protocols for multi-scale assessment (e.g., requiring methods to report not just overall accuracy but also performance on hard subsets like “dense forest canopy” vs “isolated trees”) and encourage inclusion of auxiliary checks like the allometric consistency mentioned above.
+In short, a benchmarking initiative for large-scale tree segmentation would provide an invaluable feedback loop, where challenges discovered by one team (e.g., failure cases in a certain region or condition) become part of the test set that everyone then tries to solve.
+This kind of iterative improvement cycle has been crucial in other domains of AI.
+We envision something like an annual challenge where models compete on a standard large-scale tree mapping task – pushing methods to be not only accurate but also general and validation-friendly.
+By establishing these community standards, we move toward a future where claims of model performance are transparent and believable, because they’ve been vetted on a broad, agreed-upon spectrum of scenarios.
 
-\textbf{Innovate in ground-truth data collection and labeling.} Lastly,
-none of the above model improvements obviates the need for better and
-more plentiful validation data. We therefore recommend parallel efforts
-to enhance ground-truth collection through automation and crowdsourcing.
-On the automation side, advances in high-resolution imaging and onboard
-AI mean that drones or small aircraft could be deployed to automatically
-detect and delineate trees in sample areas, providing semi-automated
-annotations that experts only need to lightly verify. For instance, a
-drone flying over a forest could use its own simpler model to suggest
-crown boundaries, which are then corrected by a human specialist --
-vastly speeding up the annotation process compared to drawing polygons
-from scratch. Similarly, leveraging crowdsourcing and citizen science
-could dramatically expand validation datasets: non-experts can be asked
-to label trees on accessible platforms (especially for easy tasks like
-clicking the center of a visible tree), and with enough redundancy and
-quality control, these contributions can rival expert labels. Projects
-for mapping trees in cities (e.g., through apps that engage the public
-to identify street trees) hint at this potential. Of course, when using
-crowd data, one must account for variability in skill and ensure
-rigorous cleaning and validation of the contributed labels -- but as a
-complementary source of truth data, it could fill gaps in areas
-professionals have not covered. Another approach is active learning,
-where the model itself guides where more data is needed: the model can
-flag areas or examples where it is most uncertain or makes conflicting
-predictions (say, an area of weirdly segmented crowns), and those areas
-would be prioritized for human labeling. This way, instead of randomly
-sampling locations for ground truth, we focus effort on the most
-informative samples -- those likely to teach the model something new or
-reveal a blind spot. Active learning strategies have been shown to
-greatly reduce the amount of data needed for training in other remote
-sensing tasks by smartly selecting the right samples to label
-{[}\^{}{]}. Integrating an active learning loop into large-scale
-validation means we continually refine the ground truth in the areas
-that matter most, thereby improving the model in a targeted fashion.
-Finally, we should not overlook the role of policy and open data
-initiatives: governments and organizations conducting tree inventories
-or LiDAR scans should be encouraged to share these as open benchmarks.
-Even if such datasets are not perfectly aligned with satellite imagery,
-they can often be matched or used to validate parts of a map. In
-summary, closing the validation data gap will require creativity --
-using machines to help label, using the crowd to scale up, and using the
-model's own intelligence to guide where to look next. These efforts
-complement the technical recommendations above: better models make use
-of the new data more effectively, and better data enables more powerful
-models.
+**Innovate in ground-truth data collection and labeling.** Lastly, none of the above model improvements obviates the need for better and more plentiful validation data.
+We therefore recommend parallel efforts to enhance ground-truth collection through automation and crowdsourcing.
+On the automation side, advances in high-resolution imaging and onboard AI mean that drones or small aircraft could be deployed to automatically detect and delineate trees in sample areas, providing semi-automated annotations that experts only need to lightly verify.
+For instance, a drone flying over a forest could use its own simpler model to suggest crown boundaries, which are then corrected by a human specialist – vastly speeding up the annotation process compared to drawing polygons from scratch.
+Similarly, leveraging crowdsourcing and citizen science could dramatically expand validation datasets: non-experts can be asked to label trees on accessible platforms (especially for easy tasks like clicking the center of a visible tree), and with enough redundancy and quality control, these contributions can rival expert labels.
+Projects for mapping trees in cities (e.g., through apps that engage the public to identify street trees) hint at this potential.
+Of course, when using crowd data, one must account for variability in skill and ensure rigorous cleaning and validation of the contributed labels – but as a complementary source of truth data, it could fill gaps in areas professionals have not covered.
+Another approach is active learning, where the model itself guides where more data is needed: the model can flag areas or examples where it is most uncertain or makes conflicting predictions (say, an area of weirdly segmented crowns), and those areas would be prioritized for human labeling.
+This way, instead of randomly sampling locations for ground truth, we focus effort on the most informative samples – those likely to teach the model something new or reveal a blind spot.
+Active learning strategies have been shown to greatly reduce the amount of data needed for training in other remote sensing tasks by smartly selecting the right samples to label \[^\].
+Integrating an active learning loop into large-scale validation means we continually refine the ground truth in the areas that matter most, thereby improving the model in a targeted fashion.
+Finally, we should not overlook the role of policy and open data initiatives: governments and organizations conducting tree inventories or LiDAR scans should be encouraged to share these as open benchmarks.
+Even if such datasets are not perfectly aligned with satellite imagery, they can often be matched or used to validate parts of a map.
+In summary, closing the validation data gap will require creativity – using machines to help label, using the crowd to scale up, and using the model’s own intelligence to guide where to look next.
+These efforts complement the technical recommendations above: better models make use of the new data more effectively, and better data enables more powerful models.
 
-By pursuing these recommendations in concert, the field can make
-significant strides toward accurate and trustworthy large-scale tree
-mapping. What we envision is an end-to-end pipeline where robust models
-(forged through self-supervised, multi-view, terrain-aware training)
-produce segmentation maps that are continuously checked against both
-classical metrics and real-world plausibility (via benchmarks and
-allometric/field validations), and where the feedback from validation
-drives further improvement of the models through active learning and
-expanded training data.
-
-\bibliographystyle{splncs04}
-\bibliography{bibliography}
-
-\end{document}
+By pursuing these recommendations in concert, the field can make significant strides toward accurate and trustworthy large-scale tree mapping.
+What we envision is an end-to-end pipeline where robust models (forged through self-supervised, multi-view, terrain-aware training) produce segmentation maps that are continuously checked against both classical metrics and real-world plausibility (via benchmarks and allometric/field validations), and where the feedback from validation drives further improvement of the models through active learning and expanded training data.

### 2025-06-10 Validation_challenges_tree_crown_segmentation/Draft1_PerspectivePaper.qmd
latest paper added, figures added, citations fixed

@@
   quarto-ecml-html: default
-author: Taimur Khan, Jasmin Krebs, Nils Nölke, Sharad Kumar Gupta, Jonathan Renkel, Caroline Arnold (order tbd)
+author:
+  - name: Taimur Khan
+    corresponding: true 
+    affiliations:
+      - name: Helmholtz Centre for Environmental Research (UFZ)
+        address: Theodor-Lieser-Str. 4, 06120 Halle (Saale)
+        country: Germany
+        url: \{taimur.khan\}\@ufz.de
+        id: south
+    url: https://www.ufz.de/index.php?en=49404
+  - name: Calvin Cordozar Broadus Jr.
+    affiliations:
+      - name: Fictional West Coast University
+        address: Long Beach CA 90840
+        country: USA
+        url: ccb@fwcu.fake
+      - name: Secondary European Affiliation
+        address: Tiergartenstr. 17
+        city: Heidelberg
+        country: Germany
+        postal-code: 69121 
+        url: lncs@springer.com
+  - name: Antwan André Patton
+    orcid: '0000-1111-2222-3333'
+    affiliations:
+      - ref: south 
 author-running: T. Khan et al.
 abstract: |
-  Advances in deep learning models have opened up new avenues for computer vision tasks, like tree crown segmentations from remote sensing images.
-As more and more models become available, the challenges of applying the models to really large spatial scales (beyond the forest/plot level) become visible.
-One of these challenges is the validation of the tree crown detection and segmentation.
-Imprecise detections/segmentations can lead to inaccuracies in downstream tasks in forestry and urban planning, creating a barrier for the application of these models to real-life use cases.
-Attributes of training data, e.g tree sizes, phenological stages, number of polygon vertices, shadows, impact the prediction accuracy of the segmentation models, hence training data should be chosen with due diligence.
-Here we identify some of the key validation challenges based on the collective experiences of not just training deep learning models, but also applying the models to very large spatial scales in Germany (federal state level).
+ Recent progress in deep learning has greatly improved the ability to extract detailed information from remote sensing images, such as identifying and outlining individual tree crowns.
+However, as these models are scaled up for use across large geographic areas—beyond individual forests or research plots—new challenges begin to emerge.
+One of the main hurdles is validating how accurately these models detect and segment tree crowns.
+Even small errors in these predictions can affect decisions in forestry management and urban planning, limiting the real-world usefulness of such models.
+Our experience working with large-scale applications across regions in Germany highlights how factors like tree size, seasonal variations, shadow effects, and the shape of training data annotations influence model performance.
+This underlines the importance of thoughtful selection and design of training datasets.
+We discuss these challenges in validation and emphasize the need for careful model evaluation when working at broader spatial scales.
 keywords: [First Keyword, Second Keyword, Another Keyword.]
@@
 
-## Introduction
+# Introduction
 
-Accurately mapping tree crowns over large areas is critical for scaling local observations to regional and global assessments of forest health, carbon stocks, and urban greenery.
+The urgency of climate change and biodiversity loss significantly heightens the importance of large-scale tree crown segmentation.
+Forests act as crucial carbon sinks and biodiversity reservoirs [@pan2024enduring], while urban trees offer vital ecosystem services[@sharma2024urban].
+To generate comprehensive assessments of forest health, carbon stocks, and urban greenery, we need to accurately map tree crowns over vast areas, including trees both within forests and outside of them.
+These trees outside forests, often found in agricultural or urban settings, must be part of our inventories, making their detection and mapping a key priority for scaling local observations to a global context.
+Large-scale segmentation enables consistent, detailed monitoring of these vital resources, informing sustainable forest management, urban planning, and policy development.
+The ambition to create national and global inventories, such as those envisioned in @tolan2024very, hinges on the reliability of such foundational segmentation data.
 In an era marked by rapid environmental change, comprehensive mapping efforts are indispensable.
 Tree crown segmentation is a cornerstone task for these objectives, as the size and structure of a tree crown—shaped by species-specific branching patterns, site conditions, and competition for light—directly influences primary production.
-A key metric in this context is the crown projection area (CPA), defined as the vertical projection of the crown onto a horizontal plane.
-Deriving CPA provides essential insights both at the individual tree level—where it provides predictions of diameter, volume, and growth rates—and at the stand level, supporting models of competition and canopy gap dynamics.
-However, inaccuracies inherent in the segmentation process, particularly when applied at large scales, can lead to biased estimations of these critical tree variables, underscoring the urgent need to address robust validation challenges in large-scale tree crown segmentation.
+
 While recent advances in deep learning offer unprecedented capabilities for automated segmentation, their robust validation at scale presents a critical, unaddressed hurdle.
 
-Among the various tree crown variables, crown spread presents unique estimation challenges, both in field measurements and, critically, when derived from remote sensing imagery.
-Its accurate derivation is not only susceptible to inaccuracies in the initial segmentation process but also relies heavily on the methods employed to extract crown spread from segmented polygons, a process fraught with its own complexities.
-Selecting an appropriate crown spread calculation method is crucial, especially for irregularly shaped tree crown polygons.
-Even with optimized calculation methods, direct matches with ground-truth data proved low (e.g., 32% in our experience), although a broader acceptable margin of error (±5 m) captured 89% of cases.
-These results immediately point to difficulties in establishing precise validation metrics.
-Furthermore, model performance exhibits size-dependent biases: trees with small crowns (\<6 m) are often overestimated, while large crowns (\>16 m) tend to be underestimated.
-These discrepancies underscore the inherent limitations of deep learning segmentation models at the extremes of the size spectrum.
-Critically, the quality of validation data profoundly influences the assessment of crown spread accuracy.
-For instance, our experience with the local tree inventory ('Baumkataster' Halle/Saale), which provides only broad crown spread ranges (5 m intervals), severely constrains the precision of validation, highlighting the need for more granular and reliable ground-truth datasets.
-
-The integration of deep learning into remote sensing has opened new ways in how we monitor and analyse ecological systems \[@zheng2024review, @zhao2023systematic\].
-Tree crown segmentation, a task central to assessing forest health, carbon stocks, and urban greenery, has especially benefited from advances in convolutional neural networks (CNNs) and transformer-based architectures.
-Approaches leveraging high-resolution imagery have demonstrated remarkable capabilities in delineating individual tree crowns, as evidenced by works such as @weinstein2019individual, @khant2025 and @freudenberg2022individual, illustrating the potential for scalable and automated tree mapping.
-The urgency of climate change and biodiversity loss further amplifies the importance of large-scale tree crown segmentation.
-Forests serve as critical carbon sinks and biodiversity reservoirs \[@pan2024enduring\], while urban trees provide essential ecosystem services \[@sharma2024urban\].
-Large-scale segmentation enables consistent, detailed monitoring of these vital resources, informing sustainable forest management, urban planning, and policy development.
-The ambition to create national and global inventories, such as those envisioned in @tolan2024very, hinges on the reliability of such foundational segmentation data.
+The integration of deep learning into remote sensing has opened new ways in how we monitor and analyse ecological systems [@zheng2024review; @zhao2023systematic].
+Tree crown segmentation has especially benefited from advances in convolutional neural networks (CNNs) and transformer-based architectures.
+Latter was used by [@JIANG2025103114], for single tree detection.
+Approaches leveraging high-resolution imagery have demonstrated remarkable capabilities in delineating individual tree crowns [@weinstein2019individual; @khant2025deeptrees; @freudenberg2022individual], illustrating the potential for scalable and automated tree mapping.
 
@@
 Foremost among these is validation—a critical yet underexplored hurdle in ensuring the reliability and generalizability of model outputs.
-As recent studies have shown, the performance of tree crown segmentation models is highly sensitive to the characteristics of the training data, such as tree size, species diversity, seasonal variations, and image quality \[@moussaid2021tree; @cong2022citrus\].
-These factors, when coupled with the complexities of real-world landscapes, underscore the need for robust and scalable validation frameworks.
+As recent studies have shown, the performance of tree crown segmentation models is highly sensitive to the characteristics of the training data, such as tree size, species diversity, seasonal variations, vitality conditions, and image quality [@moussaid2021tree; @cong2022citrus], a fact that directly reflects the profound complexity of how tree crowns appear in real-world conditions.
+To enable scalable validation frameworks, it is therefore necessary to accurately characterize and account for this complexity.
 
-This Perspective builds on our extensive experience in applying deep learning models to large-scale tree crown segmentation across Germany, addressing federal state-level landscapes.
-By synthesizing insights from model training, deployment, and validation, we highlight the nuanced challenges associated with assessing segmentation accuracy over expansive and heterogeneous terrains.
-These challenges are not mere technicalities—they bear significant implications for downstream applications in forestry, conservation, and urban planning.
-Through this work, we aim to not only highlight these critical validation hurdles but also to foster a dialogue on the methodological innovations required to ensure that deep learning advancements in remote sensing translate into actionable insights at the scales demanded by global environmental challenges.
+A key metric in this context is the crown projection area (CPA), defined as the vertical projection of the crown onto a horizontal plane.
+Deriving CPA provides essential insights both at the individual tree level—where it provides predictions of diameter, volume, and growth rates—and at the stand level, supporting models of competition and canopy gap dynamics [@shimano1997analysis; @krajicek1961crown].
+However, inaccuracies inherent in the segmentation process, particularly when applied at large scales, can lead to biased estimations of these critical tree variables, underscoring the urgent need to address robust validation challenges in large-scale tree crown segmentation.
 
-## Case Study: Tree Crown Segmentation in Saxony and Saxony-Anhalt, Germany
+Among the available crown metrics, crown spread is particularly valuable as an independent variable for validating segmentation results, as it captures essential information about tree size and structure.
+However, accurate estimation of crown spread is complex in both field and remote sensing contexts.
+Errors can arise not only during the segmentation process but also from the methods used to derive crown spread from segmented polygons, especially for irregularly shaped crowns.
+The choice of calculation method strongly influences the quality of crown spread estimates and, consequently, the reliability of validation metrics.
+Furthermore, the accuracy of validation is fundamentally dependent on the quality of reference data; coarse or inconsistent ground-truth datasets can substantially limit the precision of assessment.
 
-In this case study, the DeepTrees model \[@khan2025\] was applied in a one-shot prediction approach using pretrained model weights provided by DeepTrees, applied to high-resolution multispectral digital orthophoto imagery (DOP20) covering the German federal states of Saxony (SN) and Saxony-Anhalt (ST).
-Sachsen, covering approximately 18,450 km², and Saxony-Anhalt, spanning around 20,452 km², represent diverse ecological and urban landscapes, ideal for assessing large-scale segmentation model performance.
+Building on our extensive experience in applying deep learning models to large-scale tree crown segmentation across Germany at the federal‐state level, this paper (1) identifies and characterizes the key challenges involved in assessing segmentation accuracy over expansive, heterogeneous terrains––including variability in canopy structure, imaging conditions, and regional ecological gradients––and (2) provides concrete methodological recommendations to address these challenges.
+By coupling a rigorous evaluation of terrain‐specific segmentation limitations with these targeted recommendations, we aim to foster a dialogue on the methodological innovations required to ensure that deep learning advancements in remote sensing translate into actionable insights at the scales required by global environmental challenges.
 
-DOP20 imagery provides detailed spatial information at 20 cm resolution per pixel, capturing visual and near infrared spectral regions essential for accurate tree crown delineation.
-Using these pretrained weights, the DeepTrees model identified approximately 218.7 million individual tree crowns—137.3 million in Saxony and 81.4 million in Saxony-Anhalt.
-The resulting segmentation dataset has been made publicly available on Zenodo (https://zenodo.org/record/exampledoi), facilitating further research and validation.
+# Case Study: Tree Crown Segmentation in Saxony and Saxony-Anhalt, Germany
 
-This comprehensive segmentation highlights substantial regional variability in tree distribution, reflecting ecological, topographical, and land-use gradients.
-The case study thus forms a practical foundation for addressing the validation challenges detailed subsequently, emphasizing the necessity for robust validation frameworks to ensure reliable ecological monitoring and informed decision-making based on large-scale automated analyses.
+To evaluate the performance and limitations of large-scale crown segmentation and its validation, the DeepTrees model [@khan2025deeptrees] was applied in a one-shot prediction approach using pretrained model weights provided by Freudenberg et al [@freudenberg2022individual] as well as model weights trained in DeepTrees [@khan2025deeptrees], applied to high-resolution multispectral digital orthophoto imagery (DOP20) covering the German federal states of Saxony (SN) and Saxony-Anhalt (ST).
+SN, covering approximately 18,450 km², and Saxony-Anhalt, spanning around 20,452 km², represent diverse ecological and urban landscapes, ideal for assessing large-scale segmentation model performance (Figure 3 in Appendices).
+The 4-channel (RGBi) DOP20 imagery, with a spatial resolution of 20 cm per pixel , enables precise tree crown delineation.
+The DeepTrees model identified approximately 218.7 million individual tree crowns—137.3 million in SN and 81.4 million in ST (Figure 1a).
+The resulting segmentation dataset has been made publicly available on Zenodo (https://zenodo.org/record/exampledoi).
 
-*Figure 1: 6 panels (1 for each state). From left to right \-\> 1\) tiles with date classification 2\) Land use (CORINE) \+ isolines. 3\) the tree segmentations.*
+Despite optimized methods, exact matches between predicted and ground-truth crown spreads were low (32%), though expanding the margin of error to ±5 m increased accuracy significantly (89%), highlighting validation challenges.
+The segmentation model systematically overestimated small crowns (\<6 m) and underestimated large crowns (\>16 m), yet the crown area versus spread relationship remained stable (Figure 1b), indicating consistent segmentations.
+Validation quality proved crucial; notably, the local inventory (‘Baumkataster’ Halle/Saale) provided broad intervals (5 m), limiting validation precision but still achieving an overall IOU of \~70%.The segmented tree crowns show substantial regional variability in tree distribution, reflecting ecological, topographical, and land-use gradients (Figure 3).
+To evaluate the performance and limitations of large-scale crown segmentation and its validation, the DeepTrees model[@khan2025deeptrees] was applied in a one-shot prediction approach using pretrained model weights provided by Freudenberg et al [@freudenberg2022individual] as well as our own model weights, applied to high-resolution multispectral digital orthophoto imagery (DOP20) covering the German federal states of Saxony (SN) and Saxony-Anhalt (ST).
+SN, covering approximately 18,450 km², and ST, spanning around 20,452 km², represent diverse ecological and urban landscapes, ideal for assessing large-scale segmentation model performance (Figure 3 in Appendices).
+The 4-channel (RGBi) DOP20 imagery, with a spatial resolution of 20 cm per pixel , enables precise tree crown delineation.
+The DeepTrees model identified approximately 218.7 million individual tree crowns—137.3 million in SN and 81.4 million in ST (Figure 1a).
+The resulting segmentation dataset has been made publicly available on Zenodo (https://zenodo.org/record/exampledoi).
 
-## Challenges
+Despite optimized methods, exact matches between predicted and ground-truth crown spreads were low (32%), though expanding the margin of error to ±5 m increased accuracy significantly (89%), highlighting validation challenges.
+The segmentation model systematically overestimated small crowns (\<6 m) and underestimated large crowns (\>16 m), yet the crown area versus spread relationship remained stable (Figure 1b), indicating consistent segmentations.
+Validation quality proved crucial; notably, the local inventory (‘Baumkataster’ Halle/Saale) provided broad intervals (5 m), limiting validation precision but still achieving an overall IOU of \~70%.The segmented tree crowns show substantial regional variability in tree distribution, reflecting ecological, topographical, and land-use gradients (Figure 3).
 
-Validating large-scale tree crown segmentation models reveals a web of interrelated challenges that go beyond those encountered in small, controlled datasets.
-Seasonal phenology is a moving target: the same forest can look drastically different between leaf-on summer imagery and leaf-off winter scenes.
-Models trained on one season often struggle in another, yielding inconsistent segmentation accuracy across the year.
+![]("./figures/figure1.png")
+
+# Challenges
+
+Validating large-scale tree crown segmentation models reveals a web of interrelated challenges.
+Among these challenges is phenology, a moving target: the same forest can look drastically different between leaf-on summer imagery and leaf-off winter scenes.
+Models trained on a single phenological period often struggle in another, yielding inconsistent segmentation accuracy across the year [@sapkota2024; @liu2024].
 For example, a canopy delineation that performs well on lush summer foliage may under-segment sparse autumn crowns or miss bare branches in winter.
 Such season-driven variability not only degrades model performance but also complicates validation – a one-shot model might appear accurate in one season and fail in the next, raising questions about how and when accuracy should be assessed.
-Incorporating multi-season data during both training and validation is essential, as phenological dynamics have been shown to strongly influence model generalizability \[@moussaid2021tree; @cong2022citrus\].
+Incorporating multi-season data including the key phenological stages: leaf emergence, flowering, fruiting, autumn coloring and leaf fall, during both training and validation is essential, as phenological dynamics have been shown to strongly influence model generalizability [@moussaid2021tree; @cong2022citrus].
 
-Spatial heterogeneity of landscapes, including terrain and illumination differences, poses another major hurdle for both segmentation and its validation.
-An algorithm that segments tree crowns flawlessly in a flat, well-lit park may stumble in a shadow-drenched valley or on a steep mountainside.
+Spatial and illumination heterogeneity of landscapes, driven by terrain and illumination differences, poses a major hurdle for both segmentation and its validation.
+An algorithm that segments tree crowns flawlessly in a flat, well-lit park may strugglein a shadow-drenched valley or on a steep mountainside.
 Variations in ground elevation and slope alter the angle of solar illumination, leading to uneven lighting and shadows that can confuse models.
 In mountainous or rugged terrains, trees on north-facing slopes might appear darker or partially occluded compared to those on south-facing slopes with direct sun, even if they are the same species and healthy.
-Such effects result in site-specific performance: models often need fine-tuning or can suffer accuracy drops when moved to new topographies or sensor angles.
+Such effects result in site-specific performance: accuracy drops when moving to new topographies or sensor angles.
 Weinstein et al.
-(2020) observed this kind of cross-site performance gap, where a tree detection model trained in one region underperformed when applied to a different region’s imagery without adaptation, underscoring how terrain and context influence outcomes \[@weinstein2020cross\].
+(2020) observed this kind of cross-site performance gap, where a tree detection model trained in one region underperformed when applied to a different region’s imagery without adaptation, underscoring how terrain and context influence outcomes [@weinstein2020cross].
+Fine-tuning can increase accuracy, but requires additional data and computational resources.
 For validation, this means that accuracy estimates from one area may not transfer to another – a critical issue when assessments at national or global scales are required.
@@
 A further fundamental challenge lies in the scarcity of accurate ground-truth data at scale.
-Reliable validation hinges on high-quality reference data (the “ground truth”), yet collecting detailed crown delineations over large regions is logistically difficult and expensive.
-Field surveys can map individual trees with great precision (e.g., measuring trunks and canopy spread on the ground), but doing this over thousands of square kilometers is infeasible.
-Aerial and satellite imagery provide broader coverage for reference data, but even these often lack the resolution or fidelity to unequivocally label each tree crown for validation purposes.
+Reliable validation hinges on high-quality reference data (the "ground truth"), yet collecting detailed crown delineations over large regions is logistically difficult and expensive.
+For instance, field surveys can map individual trees with great precision, providing highly accurate tree position information (e.g., through GPS measurements of trunk location or detailed canopy spread on the ground) [@steier2024].
+However, doing this over thousands of square kilometers is infeasible.
+Conversely, while aerial and satellite imagery offer broad coverage, even at high resolutions, they present significant challenges for unequivocally labeling each tree crown for validation purposes.
+This is due to factors like canopy overlap, varied lighting conditions, complex tree morphologies, and the sheer scale of the areas to be annotated, which introduce ambiguities and make consistent manual delineation impractical or prohibitively expensive across large regions.
 UAV (drone) campaigns can bridge the gap by capturing very high-resolution images or LiDAR of sample areas, but they are limited in flight range and still require extensive human annotation to turn imagery into usable ground truth.
@@
 Compounding the issue of limited data is the inconsistency in reference annotations and evaluation metrics.
-Even when ground-reference data exist, their format can differ – sometimes reference trees are marked by a single GPS point (e.g., trunk location), sometimes by a hand-drawn polygon outlining the crown.
-This creates a challenge in validation: how do we decide if a predicted crown polygon “matches” a ground-truth point, or how to handle cases where one field-mapped tree corresponds to multiple overlapping crown segments in the image?(Figure ref)  Conversely, field crews might delineate a broad canopy as one crown while an automated model splits it into two segments (or vice versa), especially in dense stands where crowns merge.
+Even when ground-reference data exist, their format can differ – sometimes reference trees are marked by a single GPS point (e.g., trunk location), sometimes by a hand-drawn polygon outlining the single crown.
+This creates a challenge in validation: how do we decide if a predicted crown polygon “matches” a ground-truth point, or how to handle cases where one field-mapped tree corresponds to multiple overlapping crown segments in the image?
+(Figure 2c) Conversely, field crews might delineate a broad canopy as one crown while an automated model splits it into two segments (or vice versa), especially in dense stands where crowns merge.
 These ambiguities in one-to-one correspondence make it hard to define what a “correct” segmentation is.
@@
 In an extreme case, a model could slightly over-segment every tree (splitting each true crown into two smaller polygons) and still achieve a reasonable IoU, despite doubling the perceived tree count – a significant error for applications.
-On the other hand, object-centric metrics such as panoptic segmentation quality attempt to consider both detection and delineation of objects \[@kirillov2019panopticsegmentation\].
+On the other hand, object-centric metrics such as panoptic segmentation quality attempt to consider both detection and delineation of objects [@kirillov2019].
 Panoptic metrics combine aspects of object detection (was each tree detected?) with segmentation quality (was each crown correctly outlined?), which can be more appropriate for tree mapping.
@@
 There is also the issue of scale and resolution in validation reporting.
-A model’s accuracy can appear to vary depending on the spatial scale at which it’s evaluated.
-For instance, a segmentation model might achieve high overall accuracy when averaged over an entire region, yet if one zooms into a small test area (say a single city park or forest stand), the error rate might be much higher or lower.
+A model’s accuracy can appear to vary depending on the spatial scale at which it is evaluated.
+For instance, a segmentation model might achieve high overall accuracy when averaged over an entire large region, yet if one zooms into a small test area (say a single city park or forest stand), the error rate might be much higher or lower.
 This can happen if errors are not evenly distributed: the model could perform very well in one type of landscape (e.g., neat urban street trees) and poorly in another (dense natural forest), and a coarse regional average could mask these extremes.
@@
 Ensuring that validation is robust across scales is tricky – one must balance broad coverage with local detail.
-It calls for multi-scale validation approaches, where accuracy is reported at multiple grain sizes or stratified by landscape type.
-Highlighting this, one could imagine a figure plotting model accuracy as a function of spatial extent or across different habitat types, illustrating how performance can drop off in specific challenging subsets despite looking good overall.
-This emphasizes that heterogeneous performance is itself a challenge to acknowledge in validation: we need methods to detect where and why a model fails, not just an aggregate score.
+It calls for multi-scale validation approaches, where accuracy is reported at multiple grain sizes or stratified by landscape or habitat  type.
 
-To summarize, validating large-scale tree segmentation models is far from a straightforward task.
-Seasonal changes, diverse terrain and illumination, limited and inconsistent ground truth, ambiguous evaluation criteria, and scale-dependent performance all intertwine to create a demanding setting.
-These challenges are not just academic – they directly impact how much trust we can place in AI-generated tree maps for real-world decisions in forestry, ecology, and urban planning.
-Recognizing these pain points is the first step; the next is devising strategies to overcome them.
-We now turn to several recommendations aimed at improving both models and validation practices, with an eye toward bridging the gap between controlled experiments and the complexity of continental-scale deployments.
+The above challenges reflect the multifaceted difficulties of validating tree crown segmentation at scale. For an at-a-glance overview of these issues and recommendations, see Table 1 in the Appendix.
 
-*Table 1: Potential validation challenges and corresponding mitigation strategies.
-This table could list key challenges in validating large-scale tree segmentations alongside proposed approaches to address them.
-For example: (Challenge) Phenological variation between seasons – (Mitigation) use multi-temporal training data and season-specific validation sets; (Challenge) Heterogeneous terrain and shadowing – (Mitigation) integrate digital elevation models and terrain-aware modeling; (Challenge) Scarcity of ground-truth annotations – (Mitigation) leverage semi-automated labeling, crowdsourcing, and active learning to expand validation data; (Challenge) Ambiguity in crown delineation – (Mitigation) adopt object-based accuracy metrics and consensus protocols for what counts as a single tree; (Challenge) Scale-dependent performance – (Mitigation) evaluate models at multiple spatial scales and stratify results by landscape type.
-This overview would visually reinforce the narrative that each challenge has an identifiable path forward, linking the “Challenges” and “Recommendations” sections.*
+![]("./figures/figure2.png")
 
-*![][image1]*  
-*Figure 2: Incomplete*
+*Figure 2:* Illustration of key validation challenges in large-scale tree crown segmentation across two federal states (Sachsen-Anhalt; in blue & Sachsen; in green) in Germany.
+Insets highlight specific issues: (a) seasonal variation (phenology) affecting appearance, (b) tiling artifacts in image preprocessing, (c) misalignment between predicted crowns and known tree stem locations, and (d) discrepancies between predicted crowns (blue) and labeled ground truth (red), emphasizing model performance gaps.
+Blue (Sachsen-Anhalt) and green (Sachsen) overlays represent modelled individual tree crowns across different Federal States in Germany.
 
-## Recommendations
+# Recommendations
 
-Addressing the above challenges requires a multi-pronged approach, combining improvements in modeling techniques with innovations in validation methodology and data collection.
-We outline here several complementary strategies that, together, can bolster the reliability of large-scale tree segmentation efforts.
-These recommendations emphasize building more robust models through advanced training paradigms, as well as creating better frameworks to evaluate and support those models in real-world conditions.
-The overarching goal is to ensure that deep learning advancements in tree mapping translate into trustworthy, actionable insights at regional to global scales.
+Validating large-scale tree segmentation models is challenging due to seasonal variability, diverse terrain and illumination conditions, inconsistent ground truth data, ambiguous evaluation criteria, and scale-dependent performance.
+These factors significantly affect the reliability and practical value of AI-generated tree maps for forestry, ecology, and urban planning.
 
-**Leverage self-supervised learning for robust representations.** A key step towards more generalizable segmentation models is tapping into unannotated data via self-supervised learning (SSL).
-Unlike traditional supervised training which is bottlenecked by limited labeled examples, SSL allows models to learn from the abundant pool of unlabeled remote sensing images – for example, by predicting missing pieces of an image or distinguishing augmented views of the same scene.
-By pre-training a model on large geospatial datasets without any human-provided labels, the model can absorb intrinsic patterns of the landscape: textures, shapes, seasonal changes, and other context that are common across images.
-These rich foundational representations can then be fine-tuned for tree crown segmentation with far fewer annotated samples than would otherwise be needed.
-Recent efforts like PhilEO Bench demonstrate the promise of this approach, evaluating geospatial foundation models that were pre-trained using SSL techniques \[@fibaek2024phileo\].
-The results show improved performance on a range of tasks (e.g.
-building footprint extraction, road mapping) after such pre-training, compared to models trained from scratch.
-In the context of trees, a model with an SSL-pretrained backbone may already “know” about basic vegetation structures, shadows, and seasonal appearances, making it more adept at delineating crowns under varied conditions.
-For instance, a foundation model trained on year-round satellite imagery might implicitly understand the difference between a leafless oak in winter and the same oak in summer, and thus require only a light fine-tuning to accurately segment each.
-Mendieta et al.
-(2023) take this further by continually pre-training on new data distributions (a form of continual learning), which helped build a Geospatial Foundation Model (GFM) that excelled across multiple remote sensing tasks \[@mendieta2023gfm\].
-Such continual SSL training could allow segmentation models to keep improving as more unlabeled data (e.g., new satellite images over time or from new regions) become available, staying up-to-date with changes in landscapes and sensor characteristics.
-Exploiting SSL is a powerful way to tackle the twin issues of limited labels and dataset bias, yielding a model that is better equipped for the diverse scenarios encountered in large-scale mapping.
+Addressing these challenges necessitates a multi-faceted approach that integrates advanced modeling techniques, improved validation methods, and robust data collection.
+Recommended strategies include enhancing model robustness through sophisticated training approaches and developing standardized evaluation frameworks to assess real-world performance accurately.
+These efforts aim to ensure that advances in deep learning yield reliable, actionable insights for large-scale tree mapping.
 
-**Integrate multi-view and multi-temporal data for consistency.** Another promising avenue is to train and evaluate models with multi-view inputs – that is, imagery of the same trees captured from different angles, sensors, or times.
-Multi-view here can mean different things: multi-angle (such as combining nadir and oblique aerial images), multi-platform (satellite imagery plus drone photos), or multi-temporal (images taken in different seasons or years).
-By exposing models to such spatiotemporal diversity, we can help them learn invariances that make segmentation more reliable.
-For example, a tree crown seen from directly above might have a certain shape, but from a side angle the outline might blend with neighbors; a model trained with both perspectives could learn to robustly identify the crown in either view.
-Likewise, combining leaf-on and leaf-off images of the same forest during training can force a model to rely on structural features (branches, trunk hints, relative spacing) in addition to just greenness, thereby improving its ability to generalize across seasons.
-Self-supervised methods are particularly well-suited to capitalize on multi-view data because they can be set up to encourage the model to produce similar latent representations for different views of the same object.
-Techniques such as masked autoencoders or contrastive learning can be used on multi-view datasets to make a model predict or align one view from another, without any manual labels needed.
-Such approaches have already shown success: for instance, researchers have used masked image modeling and contrastive SSL on multi-view satellite imagery to boost performance on segmentation and detection tasks, essentially teaching the model that “these two different-looking images actually contain the same trees” \[@mukkavilli2023prithvi; @fibaek2024phileo\].
-By training with multi-view consistency, the model becomes more robust to viewpoint and temporal changes, which directly addresses the phenology and illumination challenges discussed earlier.
-Importantly, this not only helps the model’s predictions but also strengthens validation, because a model that is consistent across views makes it easier to compare predicted and true crowns even when the reference data comes from a slightly different perspective or date.
-In practice, one could imagine a validation scheme where the agreement of a model’s output between leaf-on and leaf-off images serves as an indicator of reliability: large discrepancies might flag areas for further human inspection.
-Overall, multi-view and multi-temporal training imbue models with a form of contextual intelligence about the 3D and time-varying nature of trees, making segmentation outcomes more stable across the real-world variability that large-scale applications inevitably encompass.
+**Leveraging self-supervised learning (SSL)** enhances the generalizability of segmentation models by utilizing abundant unlabeled remote sensing data.
+Unlike traditional supervised methods constrained by limited annotations, SSL enables models to derive meaningful representations through tasks such as predicting missing image parts or distinguishing augmented views of scenes.
+Pre-training models with extensive unlabeled geospatial imagery allows them to internalize intrinsic landscape patterns—textures, shapes, and seasonal variations—which can be efficiently fine-tuned with fewer labeled samples for specific tasks like tree crown segmentation.
+Recent research underscores the effectiveness of SSL-based geospatial foundation models.
+PhilEO Bench demonstrated improved performance across multiple remote sensing applications, such as building footprint extraction and road mapping, compared to training models from scratch [@fibaek2024phileo].
+SSL-trained models inherently recognize basic vegetation structures, shadows, and seasonal dynamics, facilitating accurate segmentation under varying conditions.
+For example, a model trained on seasonal satellite imagery implicitly distinguishes between a tree’s winter and summer appearances, significantly reducing the required fine-tuning.
+Expanding this approach, Mendieta et al.
+(2023) employed continual SSL training, integrating new data distributions to develop a robust Geospatial Foundation Model (GFM) proficient across diverse remote sensing tasks [@mendieta2023gfm].
+This continual learning paradigm ensures segmentation models remain current with evolving landscapes and sensor technologies.
+Thus, SSL effectively addresses challenges associated with limited labeled data and dataset biases, improving adaptability and performance of large-scale segmentation models.
 
-**Incorporate terrain information into the modeling pipeline**.
-As noted, uneven terrain can cause substantial variability in how trees appear in images, so bringing explicit knowledge of terrain into the segmentation process is a logical remedy.
-One recommendation is to fuse digital elevation models (DEMs) or LiDAR-derived terrain data with the imagery during model training.
-This could be as simple as providing the model with an extra channel of input encoding elevation/slope, or as complex as designing the model to separately process terrain context.
-Self-supervised pre-training can be extended here too: recent work on multisensor geospatial foundation models has shown that including elevation data in SSL (e.g., tasking the model to distinguish imagery of flat vs.
-mountainous areas) yields better feature representations for downstream tasks \[@han2024msGFM\].
-By differentiating between bare earth and above-ground structures in pre-training, the model learns to discount illumination differences that are purely due to slope and aspect, focusing instead on actual objects like trees.
-In a tree segmentation scenario, a terrain-informed model could recognize that a dark region in an image is a shaded hillside rather than a non-existent gap in canopy cover, or that an elongated shape on a steep slope is still a single tree crown albeit skewed by perspective.
-Incorporating terrain data directly addresses the challenge of spatial heterogeneity: it provides a reference frame to normalize out some variability.
-This can improve validation as well – for example, error analysis can be stratified by terrain class to ensure a model works not just on average, but on hilltops and valleys alike.
-We recommend that future segmentation models, especially for large regions with varied topography, adopt terrain-aware training strategies.
-Even if a full DEM is not available everywhere, approximating slope from the imagery or using coarse global elevation data could still offer benefits.
-Ultimately, bridging the gap between pixel appearance and real-world topography makes the model’s understanding more physical and generalizable, reducing surprises when it’s deployed on a new landscape.
+**Integrating multi-view and multi-temporal data** enhances model consistency by training with diverse imagery of the same trees captured from varying angles, sensors, or times.
+Multi-view data encompass multi-angle (nadir and oblique aerial images), multi-platform (satellite and drone imagery), and multi-temporal (images from different seasons or years) perspectives, enabling models to learn robust invariances for reliable segmentation under varying conditions.
+Combining leaf-on and leaf-off images, compels the model to recognize structural features beyond mere greenness, thus improving seasonal generalization.
+Self-supervised learning (SSL) techniques, such as masked autoencoders and contrastive learning, effectively utilize multi-view datasets by encouraging models to generate consistent representations across different views without manual labels.
+Studies utilizing masked image modeling and contrastive SSL on multi-view satellite imagery have demonstrated significant performance improvements in segmentation and detection tasks [@mukkavilli2023prithvi; @fibaek2024phileo].
+Employing multi-view consistency training directly mitigates challenges associated with phenological and illumination variations, thereby enhancing model robustness and facilitating validation.
+For example, discrepancies in model predictions between leaf-on and leaf-off imagery could flag areas requiring further examination.
+Thus, multi-view and multi-temporal training provide models with contextual understanding of the dynamic and three-dimensional nature of trees, stabilizing segmentation results across real-world variability.
 
-**Focus model attention on domain-specific features.** Beyond data augmentation and multi-source inputs, improvements in the model architecture and training objectives themselves can yield more reliable segmentation.
-One intriguing direction is the use of feature-guided masked autoencoders or similar techniques that encourage the model to learn high-level features rather than getting bogged down in pixel-level noise.
-In remote sensing, not all pixels are equal – the spectral signature of a healthy tree canopy, for instance, is characterized by certain reflectance patterns (like high near-infrared reflectance for foliage), and textural cues can differentiate a tree from grass.
-A vanilla model might or might not latch onto these subtle cues.
-However, a feature-guided approach explicitly trains the model to reconstruct or predict meaningful feature representations (such as vegetation indices or edge maps) instead of raw pixel intensity.
-For example, a recent approach called FG-MAE (Feature Guided Masked Autoencoder) masks out parts of an image and tasks the model with predicting domain-relevant features (like a NDVI – Normalized Difference Vegetation Index image, or other engineered representations) for the masked region rather than the raw pixels.
-This forces the model to infer what type of object should be there, not just to copy textures, thereby learning a more semantic understanding of the scene \[@mukkavilli2023prithvi\].
-Applying this idea to tree segmentation, we could pre-train models to predict features that highlight vegetation structure (perhaps oriented gradients that capture tree crown edges, or canopy height estimations from stereo images) so that the model’s internal representations become highly attuned to “tree-ness.” When fine-tuned for segmentation, such a model may be better at delineating tangled canopies or differentiating trees from confusing background elements, because it has learned to focus on the attributes that define a tree in imagery.
-Early studies in multispectral and SAR domains have found that this approach yields improved performance in complex environments \[@mukkavilli2023prithvi\].
-We recommend integrating feature-guidance in training for large-scale tree mapping, especially in areas with complex backgrounds (e.g., urban environments where trees mingle with buildings).
-Not only does this likely boost accuracy, it could also produce more interpretable model outputs or uncertainties – a model that knows what features it’s looking for might provide reasoning (explicit or implicit) for its segmentation, which in turn aids validation and error diagnosis.
+**Integrating terrain data into tree segmentation models** addresses image variability caused by uneven topography.
+Fusing digital elevation models (DEMs) or LiDAR-derived terrain data with imagery during model training can enhance segmentation accuracy.
+This can involve providing elevation/slope as an additional input channel or designing models to process terrain context separately.
+Self-supervised pre-training can leverage elevation data to improve feature representations.
+By differentiating between bare earth and above-ground structures, models learn to distinguish actual objects from illumination differences caused by slope and aspect.
+For instance, a terrain-informed model can differentiate shaded hillsides from canopy gaps or recognize a single tree crown on a steep slope despite perspective distortion [@han2024msGFM].
+Incorporating terrain data directly addresses spatial heterogeneity, providing a reference frame to normalize variability and improve validation.
+This allows for error analysis stratified by terrain class, ensuring consistent performance across diverse topographies.
+We recommend that future segmentation models, particularly for regions with varied topography, adopt terrain-aware training strategies.
+Even with limited DEM availability, approximating slope from imagery or using coarse global elevation data can be beneficial.
+Ultimately, integrating real-world topography with pixel appearance enhances model understanding and generalizability, leading to more robust deployment in new landscapes.
 
-**Expand validation beyond pixel agreement – use ecological consistency checks.** Traditional validation of segmentation focuses on geometric overlap with ground truth shapes, but for tree mapping we can also exploit well-established ecological relationships as an additional form of validation.
-Trees have allometric relationships – mathematical links between dimensions like trunk diameter, height, crown diameter, and biomass.
-These relationships have been measured in forestry for decades.
-For example, a tree of a given height typically has a crown of roughly proportional diameter, and there are known bounds on how big a crown can get for a given trunk size.
-We propose using such allometric equations as a sanity check for segmentation outputs.
-If a model’s predicted tree crowns violate basic allometry (say, a tiny tree height but a huge crown width, or a cluster of crowns whose total basal area implies an impossible biomass for that area), it might indicate errors in the segmentation or missing trees.
-One could quantitatively compare the distribution of predicted crown sizes and tree heights (if height data or estimates are available) against expected distributions from field data.
-Significant deviations could flag problems: for instance, if the model frequently delineates very large crowns that in reality would correspond to 80-meter tall trees (which don’t exist in the region), those are likely over-segmentation artifacts.
-Conversely, if predicted crowns are all very small in an old-growth forest where trees should be large, the model might be under-segmenting (splitting one crown into many).
-Researchers have indeed used this kind of approach in related contexts; for example, Song et al.
-(2023) employed Gaussian process regression on tree height to estimate biomass, demonstrating how linking remote sensing outputs to allometric models can validate whether the outputs make ecological sense \[@song2023biomass\].
-In practice, implementing allometric checks means bringing in additional data or models (e.g., a LiDAR-derived tree height map, or species-specific allometric formulas from forestry literature) and cross-verifying the plausibility of the AI-generated tree map.
+**Focus model attention on domain-specific features.** Advancements in model architecture and training objectives, such as feature-guided masked autoencoders, can enhance segmentation reliability by directing learning toward high-level, domain-relevant features rather than pixel-level noise.
+In remote sensing, specific spectral and textural cues (e.g., high near-infrared reflectance in healthy canopies) distinguish tree crowns from other land covers.
+Feature-guided methods train models to reconstruct meaningful feature representations—such as vegetation indices or edge maps—instead of raw pixels.
+For example, FG-MAE (Feature Guided Masked Autoencoder) tasks the model with predicting domain-specific features (e.g., NDVI or engineered representations) for masked image regions, promoting semantic understanding rather than texture replication [@mukkavilli2023prithvi].
+Applied to tree segmentation, such pre-training focuses internal representations on vegetation structure (e.g., crown edges or canopy height), improving delineation of complex canopies and separation from backgrounds.
+Early studies in multispectral and SAR imagery confirm improved segmentation in challenging environments [@allen2025manual].
+Integrating feature-guidance is thus recommended for large-scale tree mapping, particularly in complex landscapes, as it not only increases accuracy but may also yield more interpretable outputs and uncertainties to support validation and error diagnosis.
+
+**Expand validation beyond pixel agreement – use ecological consistency checks.** Traditional segmentation validation relies on geometric overlap with ground truth, but tree mapping can benefit from leveraging ecological allometric relationships as an additional check.
+Allometry—well-established links between tree dimensions such as trunk diameter, height, and crown width—provides expected bounds for tree size relationships.
+Applying allometric equations to segmentation outputs serves as a “sanity check”: for example, predictions where crown sizes and tree heights deviate significantly from field-based distributions may indicate model errors or missing trees.
+Overly large predicted crowns in a region without tall trees, or consistently small crowns in old-growth forests, can reveal segmentation artifacts.
+Recent studies, such as Song et al.
+(2023), have used statistical models to connect remote sensing outputs to allometric expectations, highlighting ecological validation as a valuable approach [@song2023biomass].
+Implementing such checks requires integrating ancillary data, such as LiDAR-derived heights or species-specific formulas, to cross-validate AI-generated tree maps.
 This recommendation shifts validation from a purely computer-vision perspective to an application-oriented perspective: after all, if the ultimate goal is to use these maps for carbon accounting, biodiversity, or forestry, then passing an ecological reality check is as important as scoring well on IoU.
@@
 **Establish community benchmarks and evaluation frameworks.** The field of geospatial AI is recognizing the value of standard benchmarks – datasets and metrics on which different methods can be compared in a reproducible way.
-For computer vision in general, ImageNet and COCO served this role, and now geospatial analogues are emerging.
-Efforts like GEO-Bench and PANGAEA have started assembling diverse geospatial tasks into comprehensive evaluation suites, underscoring the importance of benchmarking models across varying conditions \[@marsocci2024pangaea\].
-We recommend developing similar benchmark datasets and challenges specifically for tree crown segmentation and mapping.
-These benchmarks should cover a wide range of landscapes (different forest types, urban trees, plantations, etc.), seasons, and remote sensing data sources, reflecting the real-world diversity highlighted in our Challenges section.
-Importantly, a common benchmark would promote consensus on validation metrics – if the community agrees to evaluate on, say, a mix of IoU for segmentation quality and perhaps a detection metric for counting accuracy, it would standardize how results are reported.
-This comparability accelerates progress: researchers can pinpoint which innovations truly lead to better performance and robustness.
-Moreover, a shared evaluation framework can include protocols for multi-scale assessment (e.g., requiring methods to report not just overall accuracy but also performance on hard subsets like “dense forest canopy” vs “isolated trees”) and encourage inclusion of auxiliary checks like the allometric consistency mentioned above.
-In short, a benchmarking initiative for large-scale tree segmentation would provide an invaluable feedback loop, where challenges discovered by one team (e.g., failure cases in a certain region or condition) become part of the test set that everyone then tries to solve.
-This kind of iterative improvement cycle has been crucial in other domains of AI.
-We envision something like an annual challenge where models compete on a standard large-scale tree mapping task – pushing methods to be not only accurate but also general and validation-friendly.
-By establishing these community standards, we move toward a future where claims of model performance are transparent and believable, because they’ve been vetted on a broad, agreed-upon spectrum of scenarios.
+Standard datasets and metrics like ImageNet and COCO have significantly impacted computer vision, inspiring geospatial counterparts such as GEO-Bench and PANGAEA [@marsocci2024pangaea].
+We propose developing specific benchmarks for tree crown segmentation and mapping, covering diverse landscapes, seasons, and remote sensing sources.
+Standardized evaluation metrics, such as IoU for segmentation and detection accuracy, would ensure comparability and transparency in results.
+Additionally, benchmarks should require multi-scale assessments, evaluating not only overall accuracy but also performance in challenging subsets (e.g., dense forests vs.
+isolated trees), incorporating auxiliary criteria like allometric consistency.
+Regular competitive challenges using standardized large-scale tree mapping tasks can accelerate method improvement and robustness.
+Ultimately, these community-driven benchmarks facilitate transparent, credible model evaluations and iterative progress in geospatial AI.
 
-**Innovate in ground-truth data collection and labeling.** Lastly, none of the above model improvements obviates the need for better and more plentiful validation data.
-We therefore recommend parallel efforts to enhance ground-truth collection through automation and crowdsourcing.
-On the automation side, advances in high-resolution imaging and onboard AI mean that drones or small aircraft could be deployed to automatically detect and delineate trees in sample areas, providing semi-automated annotations that experts only need to lightly verify.
-For instance, a drone flying over a forest could use its own simpler model to suggest crown boundaries, which are then corrected by a human specialist – vastly speeding up the annotation process compared to drawing polygons from scratch.
-Similarly, leveraging crowdsourcing and citizen science could dramatically expand validation datasets: non-experts can be asked to label trees on accessible platforms (especially for easy tasks like clicking the center of a visible tree), and with enough redundancy and quality control, these contributions can rival expert labels.
-Projects for mapping trees in cities (e.g., through apps that engage the public to identify street trees) hint at this potential.
-Of course, when using crowd data, one must account for variability in skill and ensure rigorous cleaning and validation of the contributed labels – but as a complementary source of truth data, it could fill gaps in areas professionals have not covered.
-Another approach is active learning, where the model itself guides where more data is needed: the model can flag areas or examples where it is most uncertain or makes conflicting predictions (say, an area of weirdly segmented crowns), and those areas would be prioritized for human labeling.
-This way, instead of randomly sampling locations for ground truth, we focus effort on the most informative samples – those likely to teach the model something new or reveal a blind spot.
-Active learning strategies have been shown to greatly reduce the amount of data needed for training in other remote sensing tasks by smartly selecting the right samples to label \[^\].
-Integrating an active learning loop into large-scale validation means we continually refine the ground truth in the areas that matter most, thereby improving the model in a targeted fashion.
-Finally, we should not overlook the role of policy and open data initiatives: governments and organizations conducting tree inventories or LiDAR scans should be encouraged to share these as open benchmarks.
-Even if such datasets are not perfectly aligned with satellite imagery, they can often be matched or used to validate parts of a map.
-In summary, closing the validation data gap will require creativity – using machines to help label, using the crowd to scale up, and using the model’s own intelligence to guide where to look next.
-These efforts complement the technical recommendations above: better models make use of the new data more effectively, and better data enables more powerful models.
+**Innovate in ground-truth data collection and labeling.** Advances in modeling alone cannot eliminate the need for enhanced validation datasets.
+We recommend expanding ground-truth collection through automation, crowdsourcing, and active learning.
+Automated techniques using drones or AI-equipped aircraft can rapidly generate tree crown annotations, minimizing expert involvement.
+Crowdsourcing allows non-experts to efficiently label straightforward cases, supported by redundancy and quality controls, proven effective in urban tree mapping.
+Active learning strategies prioritize uncertain or conflicting model predictions for expert review, significantly optimizing annotation efforts [@Wu_2022_CVPR].
+Additionally, promoting open data practices and sharing existing inventories and LiDAR datasets from public agencies can provide critical validation resources.
+Integrating these approaches—automation, crowdsourcing, active learning, and open data—will effectively bridge validation data gaps and enhance model robustness and generalizability.
 
-By pursuing these recommendations in concert, the field can make significant strides toward accurate and trustworthy large-scale tree mapping.
-What we envision is an end-to-end pipeline where robust models (forged through self-supervised, multi-view, terrain-aware training) produce segmentation maps that are continuously checked against both classical metrics and real-world plausibility (via benchmarks and allometric/field validations), and where the feedback from validation drives further improvement of the models through active learning and expanded training data.
+# Conclusion
+@
+Deep learning has opened a new frontier for tree crown segmentation from remote sensing imagery, yet its promise is contingent on our ability to validate these models reliably and at scale.
+As this paper has outlined, the challenges are not merely technical—they are epistemological.
+They force us to ask: what does it mean to “know” a tree from above, when the canopy is a moving target shaped by phenology, terrain, land cover and image artifacts?
+
+The comprehensive case study in Sachsen and Sachsen-Anhalt exemplifies how substantial regional variability in ecology, topography, and land use affects model performance.
+Validation challenges identified—including phenological dynamics, spatial heterogeneity, and scale-dependent accuracy—highlight the need for robust validation frameworks tailored to large spatial extents.
+
+Emerging methods such as self-supervised learning, geospatial foundation models, and multi-view fusion offer a compelling pathway forward.
+These approaches not only reduce reliance on costly annotations but also capture the underlying structure of complex and dynamic landscapes.
+Equally, validation must evolve beyond static benchmarks.
+Indirect metrics—like allometric plausibility checks— must be brought into the fold.
+Validation, in this context, becomes less about binary correctness and more about probabilistic trust.
+
+Pursuing the recommendations, the field can significantly advance accurate and trustworthy large-scale tree mapping by integrating robust, self-supervised, multi-view, and terrain-aware models with continuous validation against classical metrics and real-world plausibility, using feedback for active learning and expanded training data.
+The real frontier is integration: aligning spatial, temporal, and ecological knowledge through a fusion of data-driven and domain-aware models.
+As these systems are deployed across continents, cities, and seasons, the imperative is not just to scale algorithms, but to scale insight.
+Only then can tree segmentation models become dependable instruments for managing the living infrastructure of our planet.
+
+# Appendices
+
+![]("./figures/figure3.png") 
+*Figure 3:*
+
+*Table 1: Summary of key validation challenges in large-scale tree crown segmentation, with examples, relevant references, and recommended mitigation strategies.*
+
+| Challenge | Description | Example/ Implications | References | Recommendation | References |
+| ----- | ----- | ----- | ----- | ----- | ----- |
+| Seasonal Phenology Variability | Tree appearance changes drastically across seasons (leaf-on summer vs.
+leaf-off winter), causing models trained in one phenological period to underperform in another.
+This leads to inconsistent segmentation accuracy and complicates validation, as a model may appear accurate in one season but fail in another.
+Multi-season data covering key phenological stages is essential for generalizable models.
+| A canopy delineation that performs well on lush summer foliage may under-segment sparse autumn crowns or miss bare branches in winter | [@sapkota2024; @liu2024; @moussaid2021tree; @cong2022citrus] | Leverage self-supervised learning (SSL) and integrate multi-view/ multi-temporal data.
+Use SSL to pre-train models on diverse, unlabeled imagery, and incorporate multi-view/multi-season data to build invariance to phenological and illumination changes.
+This enhances generalizability and reduces the need for extensive labeled data.
+| [@fibaek2024phileo; @mendieta2023gfm; @mukkavilli2023prithvi] |
+| Spatial and Illumination Heterogeneity | Differences in terrain (elevation, slope) and illumination (shadows, sun angle) affect both segmentation and validation.
+Models may perform well in flat, well-lit areas but poorly in shadowed valleys or steep slopes, resulting in site-specific performance and complicating transferability and validation at larger scales.
+| A model that segments tree crowns flawlessly in a flat, well-lit park may stumble in a shadow-drenched valley or on a steep mountainside.
+| [@weinstein2020cross] | Integrate terrain data and multi-view consistency.
+Fuse digital elevation models (DEMs) or LiDAR with imagery to provide terrain context, and train models for multi-view consistency to handle illumination and topographic variability.
+This improves robustness and allows for stratified validation by terrain class.
+| [@han2024msGFM; @mukkavilli2023prithvi; @fibaek2024phileo] |
+| Scarcity of Accurate Ground Truth Data | Collecting detailed crown delineations over large regions is logistically difficult and expensive.
+Field surveys are precise but not scalable; aerial/satellite imagery lacks resolution for unequivocal labeling due to factors like canopy overlap, varied lighting conditions, complex tree morphologies, and the sheer scale of the areas to be annotated.
+This leads to a mismatch between model output scale and validation data, introducing uncertainty and bias.
+| Ground truth typically covers only small plots, making it difficult to validate national-scale maps.
+| [@steier2024] | Innovate in ground-truth data collection and labeling.
+Combining automation (e.g., AI-equipped drones), crowdsourcing, and active learning to expand and diversify validation datasets.
+Open data sharing and targeted expert annotation can further close the validation gap.
+|   |
+| Inconsistency in Reference Annotations and Evaluation Metrics | Reference data formats vary (GPS points vs.
+polygons), leading to ambiguity in matching predictions to ground truth.
+Dense stands and overlapping crowns complicate one-to-one correspondence.
+Pixel-wise metrics (e.g., IoU) may not reflect ecological reality, while object-centric metrics (e.g., panoptic segmentation) require well-defined ground-truth objects.
+Lack of consensus on evaluation protocols results in incompatible accuracy reports across studies.
+| A model could over-segment every tree and still achieve a good IoU, but double the perceived tree count which is problematic for applications requiring accurate counts such as for managed forest or botanical gardens.
+| [@kirillov2019; @stewart2021] | Establish community benchmarks and evaluation frameworks.
+Develop standardized datasets, protocols, and metrics for tree segmentation.
+Community benchmarks should cover diverse landscapes and seasons, and include multi-scale and ecological consistency checks for robust, comparable validation.
+| [@marsocci2024pangaea; @fibaek2024phileo] |
+| Scale and Resolution Dependence in Validation Reporting | Model accuracy varies with the spatial scale of evaluation.
+Regional averages can mask local errors; performance may be high overall but poor in specific challenging areas.
+Errors may be unevenly distributed across landscape types, making it necessary to use multi-scale validation approaches and stratified reporting.
+| A model might perform well regionally but poorly in a specific forest stand or urban park, misleading local users.
+|  | Expand validation beyond pixel agreement using ecological consistency checks and multi-scale reporting.
+Integrate allometric relationships and stratified accuracy reporting to ensure ecological plausibility and transparency at multiple spatial scales.
+| [@song2023biomass; @marsocci2024pangaea] |
+
+\begin{credits}
+\subsubsection{\ackname} DeepTrees is part of the [“DeepTrees: Deep-Learning based spatiotemporal tree inventorying and monitoring from public orthoimages”](https://deeptrees.de) project, funded by the Integration Platform "Sustainable Future Land Use" at Helmholtz-Centre for Environmental Research – UFZ within the Programme oriented Funding (PoF) period IV of the Helmholtz Program "Changing Earth – Sustaining our Future", Topic 5 "Landscapes of the Future.
+This repository is based on the work described in Freudenberg et al.
+(2022).
+This work was supported by Helmholtz Association's Initiative and Networking Fund through Helmholtz AI [grant number: ZT-I-PF-5-01].
+This work used resources of the Deutsches Klimarechenzentrum (DKRZ) granted by its Scientific Steering Committee (WLA) under project ID AIM.
+
+\subsubsection{\discintname}
+The authors have no competing interests to declare that are
+relevant to the content of this article. 
+\end{credits}

### 2025-06-10 Validation_challenges_tree_crown_segmentation/Draft1_PerspectivePaper.qmd
paper done

@@
 title-running: Validation challenges in tree crown segmentations
+bibliography: bibliography.bib
 format:
@@
   - name: Taimur Khan
+    orcid: '0000-0001-7833-5474'
     corresponding: true 
@@
       - name: Helmholtz Centre for Environmental Research (UFZ)
-        address: Theodor-Lieser-Str. 4, 06120 Halle (Saale)
+        address: Theodor-Lieser-Str. 4
+        city: Halle (Saale)
+        postal-code: 04109 
         country: Germany
-        url: \{taimur.khan\}\@ufz.de
+        url: taimur.khan@ufz.de
         id: south
     url: https://www.ufz.de/index.php?en=49404
-  - name: Calvin Cordozar Broadus Jr.
+  - name: Jasmin Krebs
+    orcid: '0009-0003-9330-1231'
     affiliations:
-      - name: Fictional West Coast University
-        address: Long Beach CA 90840
-        country: USA
-        url: ccb@fwcu.fake
+      - name: Leipzig University
+        address: Augustusplatz 10 
+        city: Leipzig
+        country: Germany
+        postal-code: 04109 
+        url: jassi.krebs@gmx.de
+  - name: Sharad Kumar Gupta
+    orcid: '0000-1111-2222-3333'
+    affiliations:
+      - ref: south
       - name: Secondary European Affiliation
@@
         postal-code: 69121 
-        url: lncs@springer.com
-  - name: Antwan André Patton
+        url: sharad.gupta@ufz.de
+  - name: Jonathan Renkel
     orcid: '0000-1111-2222-3333'
     affiliations:
-      - ref: south 
+      - name: Martin Luther University Halle-Wittenberg
+        address: Von-Seckendorff-Platz 2
+        city: Halle (Saale)
+        country: Germany
+        postal-code: 06120
+        url: jonathan.renkel@geo.uni-halle.de
+  - name: Caroline Arnold
+    orcid: '0000-0002-9458-1517'
+    affiliations:
+      - name: Helmholtz-Zentrum Hereon
+        address: Max-Planck-Straße 1
+        city: Geesthacht
+        country: Germany
+        postal-code: 21502 
+        url: caroline.arnold@hereon.de
+  - name: Nils Nölke 
+    orcid: '0000-0003-4925-2287'
+    affiliations:
+      - name: University of Göttingen, Forest Inventory and Remote Sensing
+        address: Büsgenweg 5
+        city: Göttingen
+        country: Germany
+        postal-code: 37077 
+        url: nils.noelke@forst.uni-goettingen.de
+
 author-running: T. Khan et al.
 abstract: |
- Recent progress in deep learning has greatly improved the ability to extract detailed information from remote sensing images, such as identifying and outlining individual tree crowns.
-However, as these models are scaled up for use across large geographic areas—beyond individual forests or research plots—new challenges begin to emerge.
-One of the main hurdles is validating how accurately these models detect and segment tree crowns.
-Even small errors in these predictions can affect decisions in forestry management and urban planning, limiting the real-world usefulness of such models.
-Our experience working with large-scale applications across regions in Germany highlights how factors like tree size, seasonal variations, shadow effects, and the shape of training data annotations influence model performance.
-This underlines the importance of thoughtful selection and design of training datasets.
-We discuss these challenges in validation and emphasize the need for careful model evaluation when working at broader spatial scales.
-keywords: [First Keyword, Second Keyword, Another Keyword.]
+  Recent advancements in deep learning have significantly enhanced the extraction of detailed ecological information from multispectral (RGBi) aerial imagery, notably the identification and segmentation of individual tree crowns.
+Using data from our extensive case study mapping approximately 218.7 million trees across the German federal states of Sachsen and Sachsen-Anhalt, we illustrate how scaling these models beyond controlled settings introduces critical validation challenges.
+Even minor inaccuracies in tree crown segmentation can substantially impact practical applications in forestry management and urban planning.
+Our findings emphasize the influence of tree allometry, seasonal variability, shadow effects, and the characteristics of training dataset annotations on model accuracy.
+Addressing these issues necessitates careful selection and design of training data and exploration of task-specific pre-training through Foundation Models.
+We highlight these validation complexities and recommend for rigorous steps to ensure the reliability and practical utility of large-scale deep learning models in ecological and urban management contexts.
+keywords: [Deep Learning, Ecology, Forestry, Remote Sensing.]
 ---
@@
 Latter was used by [@JIANG2025103114], for single tree detection.
-Approaches leveraging high-resolution imagery have demonstrated remarkable capabilities in delineating individual tree crowns [@weinstein2019individual; @khant2025deeptrees; @freudenberg2022individual], illustrating the potential for scalable and automated tree mapping.
+Approaches leveraging high-resolution imagery have demonstrated remarkable capabilities in delineating individual tree crowns [@weinstein2019individual; @khan2025deeptrees; @freudenberg2022individual, @JIANG2025103114], illustrating the potential for scalable and automated tree mapping.
 
 However, while these deep learning models demonstrate impressive performance in controlled environments or with limited-scale datasets, their deployment at expansive spatial extents, such as regional or national scales, exposes a distinct spectrum of challenges that remain largely unaddressed.
-Foremost among these is validation—a critical yet underexplored hurdle in ensuring the reliability and generalizability of model outputs.
+Foremost among these is validation — a critical yet underexplored hurdle in ensuring the reliability and generalizability of model outputs.
 As recent studies have shown, the performance of tree crown segmentation models is highly sensitive to the characteristics of the training data, such as tree size, species diversity, seasonal variations, vitality conditions, and image quality [@moussaid2021tree; @cong2022citrus], a fact that directly reflects the profound complexity of how tree crowns appear in real-world conditions.
@@
 
-![]("./figures/figure1.png")
+![a.
+Spatial distribution of individual tree crown segmentation polygons derived from multispectral DOP20 imagery across the German federal states of Sachsen (SN, green) and Sachsen-Anhalt (ST, blue), totaling 218.7 million predicted crowns.
+b.
+Relationship between normalized crown area (x-axis) and estimated crown spread (y-axis) for all segmented trees.
+A clear positive correlation is observed, although deviations from the expected linear trend highlight segmentation uncertainties, particularly for small and irregularly shaped crowns.
+This variation underscores the need for improved methods to capture crown geometry and supports the use of allometric checks for quality assessment.]("./figures/figure1.png")
 
@@
 
-![]("./figures/figure2.png")
-
-*Figure 2:* Illustration of key validation challenges in large-scale tree crown segmentation across two federal states (Sachsen-Anhalt; in blue & Sachsen; in green) in Germany.
+![Illustration of key validation challenges in large-scale tree crown segmentation across two federal states (Sachsen-Anhalt; in blue & Sachsen; in green) in Germany.
 Insets highlight specific issues: (a) seasonal variation (phenology) affecting appearance, (b) tiling artifacts in image preprocessing, (c) misalignment between predicted crowns and known tree stem locations, and (d) discrepancies between predicted crowns (blue) and labeled ground truth (red), emphasizing model performance gaps.
-Blue (Sachsen-Anhalt) and green (Sachsen) overlays represent modelled individual tree crowns across different Federal States in Germany.
+Blue (Sachsen-Anhalt) and green (Sachsen) overlays represent modelled individual tree crowns across different Federal States in Germany.]("./figures/figure2.png")
 
@@
 # Conclusion
-@
+
 Deep learning has opened a new frontier for tree crown segmentation from remote sensing imagery, yet its promise is contingent on our ability to validate these models reliably and at scale.
@@
 
-![]("./figures/figure3.png") 
-*Figure 3:*
-
-*Table 1: Summary of key validation challenges in large-scale tree crown segmentation, with examples, relevant references, and recommended mitigation strategies.*
+![Environmental heterogeneity across the German federal states of Sachsen and Sachsen-Anhalt, where tree crown segmentation was conducted.
+(Top) Acquisition months of DOP20 imagery (April to September) used in the segmentation, showing strong spatial variation in image phenology ( [@geosn_dop20_2022]; [@lvermgeo2020dop]).
+(Middle) Aggregated land cover classes based on CORINE Land Cover 2020 data, grouped into five categories: agricultural areas, artificial surfaces, forest and semi-natural areas, water bodies, and wetlands ( [@eea_corine2020]).
+(Bottom) Elevation map derived from the COPDEM GLO-30 digital elevation model at 30 m spatial resolution, representing terrain variability used to assess segmentation accuracy across topographic gradients ( [@copernicus_copdem30]).]("./figures/figure3hor.png")
 
@@
 
+: Summary of key validation challenges in large-scale tree crown segmentation, with examples, relevant references, and recommended mitigation strategies.
+{#tbl-headings .tbl-colwidths="15,25,20,10,20,10"}
+
 \begin{credits}

### 2025-06-11 Validation_challenges_tree_crown_segmentation/Draft1_PerspectivePaper.qmd
Update Draft1_PerspectivePaper.qmd
remove duplicate paragraph
Removed a duplicate paragraph on page 4.

@@
         postal-code: 04109 
-        url: jassi.krebs@gmx.de
+        url: jk21byxu@studserv.uni-leipzig.de
   - name: Sharad Kumar Gupta
@@
 The resulting segmentation dataset has been made publicly available on Zenodo (https://zenodo.org/record/exampledoi).
-
-Despite optimized methods, exact matches between predicted and ground-truth crown spreads were low (32%), though expanding the margin of error to ±5 m increased accuracy significantly (89%), highlighting validation challenges.
-The segmentation model systematically overestimated small crowns (\<6 m) and underestimated large crowns (\>16 m), yet the crown area versus spread relationship remained stable (Figure 1b), indicating consistent segmentations.
-Validation quality proved crucial; notably, the local inventory (‘Baumkataster’ Halle/Saale) provided broad intervals (5 m), limiting validation precision but still achieving an overall IOU of \~70%.The segmented tree crowns show substantial regional variability in tree distribution, reflecting ecological, topographical, and land-use gradients (Figure 3).
 
@@
 
-# Appendices
+\begin{credits}
+\subsubsection{\ackname} DeepTrees is part of the [“DeepTrees: Deep-Learning based spatiotemporal tree inventorying and monitoring from public orthoimages”](https://deeptrees.de) project, funded by the Integration Platform "Sustainable Future Land Use" at Helmholtz-Centre for Environmental Research – UFZ within the Programme oriented Funding (PoF) period IV of the Helmholtz Program "Changing Earth – Sustaining our Future", Topic 5 "Landscapes of the Future.
+This repository is based on the work described in Freudenberg et al.
+(2022).
+This work was supported by Helmholtz Association's Initiative and Networking Fund through Helmholtz AI [grant number: ZT-I-PF-5-01].
+This work used resources of the Deutsches Klimarechenzentrum (DKRZ) granted by its Scientific Steering Committee (WLA) under project ID AIM.
+
+\subsubsection{\discintname}
+The authors have no competing interests to declare that are
+relevant to the content of this article. 
+\end{credits}
+
+<!--# Appendix
 
@@
 : Summary of key validation challenges in large-scale tree crown segmentation, with examples, relevant references, and recommended mitigation strategies.
-{#tbl-headings .tbl-colwidths="15,25,20,10,20,10"}
-
-\begin{credits}
-\subsubsection{\ackname} DeepTrees is part of the [“DeepTrees: Deep-Learning based spatiotemporal tree inventorying and monitoring from public orthoimages”](https://deeptrees.de) project, funded by the Integration Platform "Sustainable Future Land Use" at Helmholtz-Centre for Environmental Research – UFZ within the Programme oriented Funding (PoF) period IV of the Helmholtz Program "Changing Earth – Sustaining our Future", Topic 5 "Landscapes of the Future.
-This repository is based on the work described in Freudenberg et al.
-(2022).
-This work was supported by Helmholtz Association's Initiative and Networking Fund through Helmholtz AI [grant number: ZT-I-PF-5-01].
-This work used resources of the Deutsches Klimarechenzentrum (DKRZ) granted by its Scientific Steering Committee (WLA) under project ID AIM.
-
-\subsubsection{\discintname}
-The authors have no competing interests to declare that are
-relevant to the content of this article. 
-\end{credits}
+{#tbl-headings .tbl-colwidths="15,25,20,10,20,10"}-->

### 2025-06-11 Validation_challenges_tree_crown_segmentation/Draft1_PerspectivePaper.qmd
fix gaps

@@
 
-Building on our extensive experience in applying deep learning models to large-scale tree crown segmentation across Germany at the federal‐state level, this paper (1) identifies and characterizes the key challenges involved in assessing segmentation accuracy over expansive, heterogeneous terrains––including variability in canopy structure, imaging conditions, and regional ecological gradients––and (2) provides concrete methodological recommendations to address these challenges.
+Building on our extensive experience in applying deep learning models to large-scale tree crown segmentation across Germany at the federal‐state level, this paper (1) identifies and characterizes the key challenges involved in assessing segmentation accuracy over expansive, heterogeneous terrains–including variability in canopy structure, imaging conditions, and regional ecological gradients–and (2) provides concrete methodological recommendations to address these challenges.
 By coupling a rigorous evaluation of terrain‐specific segmentation limitations with these targeted recommendations, we aim to foster a dialogue on the methodological innovations required to ensure that deep learning advancements in remote sensing translate into actionable insights at the scales required by global environmental challenges.
@@
 
-![a.
+::: {.column width="100%"}
+```{=latex}
+\begin{figure}[htbp]
+\centering
+\includegraphics[width=0.8\textwidth]{figures/figure1.png}
+\caption{a.
 Spatial distribution of individual tree crown segmentation polygons derived from multispectral DOP20 imagery across the German federal states of Sachsen (SN, green) and Sachsen-Anhalt (ST, blue), totaling 218.7 million predicted crowns.
@@
 A clear positive correlation is observed, although deviations from the expected linear trend highlight segmentation uncertainties, particularly for small and irregularly shaped crowns.
-This variation underscores the need for improved methods to capture crown geometry and supports the use of allometric checks for quality assessment.]("./figures/figure1.png")
+This variation underscores the need for improved methods to capture crown geometry and supports the use of allometric checks for quality assessment.}
+\end{figure}
 
@@
 
-![Illustration of key validation challenges in large-scale tree crown segmentation across two federal states (Sachsen-Anhalt; in blue & Sachsen; in green) in Germany.
+::: {.column width="100%"}
+```{=latex}
+\begin{figure}[htbp]
+\centering
+\includegraphics[width=0.8\textwidth]{figures/figure2.png}
+\caption{Illustration of key validation challenges in large-scale tree crown segmentation across two federal states (Sachsen-Anhalt; in blue I\& Sachsen; in green) in Germany.
 Insets highlight specific issues: (a) seasonal variation (phenology) affecting appearance, (b) tiling artifacts in image preprocessing, (c) misalignment between predicted crowns and known tree stem locations, and (d) discrepancies between predicted crowns (blue) and labeled ground truth (red), emphasizing model performance gaps.
-Blue (Sachsen-Anhalt) and green (Sachsen) overlays represent modelled individual tree crowns across different Federal States in Germany.]("./figures/figure2.png")
+Blue (Sachsen-Anhalt) and green (Sachsen) overlays represent modelled individual tree crowns across different Federal States in Germany.}
+\end{figure}
 

### 2025-06-11 Validation_challenges_tree_crown_segmentation/Draft1_PerspectivePaper.qmd
fix typo

@@
 \includegraphics[width=0.8\textwidth]{figures/figure2.png}
-\caption{Illustration of key validation challenges in large-scale tree crown segmentation across two federal states (Sachsen-Anhalt; in blue I\& Sachsen; in green) in Germany.
+\caption{Illustration of key validation challenges in large-scale tree crown segmentation across two federal states (Sachsen-Anhalt; in blue \& Sachsen; in green) in Germany.
 Insets highlight specific issues: (a) seasonal variation (phenology) affecting appearance, (b) tiling artifacts in image preprocessing, (c) misalignment between predicted crowns and known tree stem locations, and (d) discrepancies between predicted crowns (blue) and labeled ground truth (red), emphasizing model performance gaps.

### 2025-06-11 Validation_challenges_tree_crown_segmentation/Draft1_PerspectivePaper.qmd
slight updates

--- 2025-06-11 Validation_challenges_tree_crown_segmentation/Draft1_PerspectivePaper.qmd
@@
         postal-code: 04109 
-        url: jk21byxu@studserv.uni-leipzig.de
+        url: jassi.krebs@gmx.de
   - name: Sharad Kumar Gupta
@@
 Using data from our extensive case study mapping approximately 218.7 million trees across the German federal states of Sachsen and Sachsen-Anhalt, we illustrate how scaling these models beyond controlled settings introduces critical validation challenges.
-Even minor inaccuracies in tree crown segmentation can substantially impact practical applications in forestry management and urban planning.
+Even minor inaccuracies in tree crown segmentation can substantially impact some practical applications in forestry management and urban planning.
 Our findings emphasize the influence of tree allometry, seasonal variability, shadow effects, and the characteristics of training dataset annotations on model accuracy.
@@
 The DeepTrees model identified approximately 218.7 million individual tree crowns—137.3 million in SN and 81.4 million in ST (Figure 1a).
-The resulting segmentation dataset has been made publicly available on Zenodo (https://zenodo.org/record/exampledoi).
+<!--The resulting segmentation dataset has been made publicly available on Zenodo (https://zenodo.org/record/exampledoi).-->
 
@@
 The DeepTrees model identified approximately 218.7 million individual tree crowns—137.3 million in SN and 81.4 million in ST (Figure 1a).
-The resulting segmentation dataset has been made publicly available on Zenodo (https://zenodo.org/record/exampledoi).
 
@@
 A clear positive correlation is observed, although deviations from the expected linear trend highlight segmentation uncertainties, particularly for small and irregularly shaped crowns.
-This variation underscores the need for improved methods to capture crown geometry and supports the use of allometric checks for quality assessment.]("./figures/figure1.png")
+This variation underscores the need for improved methods to capture crown geometry and supports the use of allometric checks for quality assessment.](./figures/figure1.png)
 
@@
 UAV (drone) campaigns can bridge the gap by capturing very high-resolution images or LiDAR of sample areas, but they are limited in flight range and still require extensive human annotation to turn imagery into usable ground truth.
-The net result is a mismatch of scales: our models aspire to map every tree across entire countries, but our ground truth typically covers only small plots or scattered samples \[^\].
+The net result is a mismatch of scales: our models aspire to map every tree across entire countries, but our ground truth typically covers only small plots or scattered samples \[\^\].
 This mismatch means that validating a “wall-to-wall” tree map often involves extrapolating from a tiny fraction of ground-referenced trees, introducing uncertainty.
@@
 Insets highlight specific issues: (a) seasonal variation (phenology) affecting appearance, (b) tiling artifacts in image preprocessing, (c) misalignment between predicted crowns and known tree stem locations, and (d) discrepancies between predicted crowns (blue) and labeled ground truth (red), emphasizing model performance gaps.
-Blue (Sachsen-Anhalt) and green (Sachsen) overlays represent modelled individual tree crowns across different Federal States in Germany.]("./figures/figure2.png")
+Blue (Sachsen-Anhalt) and green (Sachsen) overlays represent modelled individual tree crowns across different Federal States in Germany.](%22./figures/figure2.png%22)
 

### 2025-06-11 Validation_challenges_tree_crown_segmentation/Draft1_PerspectivePaper.qmd
final

@@
 bibliography: bibliography.bib
+no-links: false
 format:
@@
         id: south
-    url: https://www.ufz.de/index.php?en=49404
   - name: Jasmin Krebs
@@
   - name: Sharad Kumar Gupta
-    orcid: '0000-1111-2222-3333'
+    orcid: '0000-0003-3444-1333'
     affiliations:
       - ref: south
-      - name: Secondary European Affiliation
-        address: Tiergartenstr. 17
-        city: Heidelberg
+      - name: HZDR - Centre for Advanced System Understanding (CASUS)
+        address: Untermarkt 20
+        city: Görlitz
         country: Germany
-        postal-code: 69121 
+        postal-code: 02826 
         url: sharad.gupta@ufz.de
   - name: Jonathan Renkel
-    orcid: '0000-1111-2222-3333'
+    orcid: '0009-0002-1513-9497'
     affiliations:
       - name: Martin Luther University Halle-Wittenberg
-        address: Von-Seckendorff-Platz 2
+        address: Von-Seckendorff-Platz 4
         city: Halle (Saale)
@@
 
-To evaluate the performance and limitations of large-scale crown segmentation and its validation, the DeepTrees model [@khan2025deeptrees] was applied in a one-shot prediction approach using pretrained model weights provided by Freudenberg et al [@freudenberg2022individual] as well as model weights trained in DeepTrees [@khan2025deeptrees], applied to high-resolution multispectral digital orthophoto imagery (DOP20) covering the German federal states of Saxony (SN) and Saxony-Anhalt (ST).
+To evaluate the performance and limitations of large-scale crown segmentation and its validation, the DeepTrees model [@khan2025deeptrees] was applied in a one-shot prediction approach using pretrained model weights provided by Freudenberg et al [@freudenberg2022individual] as well as model weights trained in DeepTrees [@khan2025deeptrees], applied to high-resolution multispectral digital orthophoto imagery (DOP20) covering the German federal states of Saxony (SN) and Saxony-Anhalt (ST) [@lvermgeo2020dop; @geosn_dop20_2022].
 SN, covering approximately 18,450 km², and Saxony-Anhalt, spanning around 20,452 km², represent diverse ecological and urban landscapes, ideal for assessing large-scale segmentation model performance (Figure 3 in Appendices).
@@
 The DeepTrees model identified approximately 218.7 million individual tree crowns—137.3 million in SN and 81.4 million in ST (Figure 1a).
-<!--The resulting segmentation dataset has been made publicly available on Zenodo (https://zenodo.org/record/exampledoi).-->
+The resulting segmentation dataset has been made available upon request on Zenodo (https://doi.org/10.5281/zenodo.15638573).
 
@@
 To evaluate the performance and limitations of large-scale crown segmentation and its validation, the DeepTrees model[@khan2025deeptrees] was applied in a one-shot prediction approach using pretrained model weights provided by Freudenberg et al [@freudenberg2022individual] as well as our own model weights, applied to high-resolution multispectral digital orthophoto imagery (DOP20) covering the German federal states of Saxony (SN) and Saxony-Anhalt (ST).
-SN, covering approximately 18,450 km², and ST, spanning around 20,452 km², represent diverse ecological and urban landscapes, ideal for assessing large-scale segmentation model performance (Figure 3 in Appendices).
-The 4-channel (RGBi) DOP20 imagery, with a spatial resolution of 20 cm per pixel , enables precise tree crown delineation.
-The DeepTrees model identified approximately 218.7 million individual tree crowns—137.3 million in SN and 81.4 million in ST (Figure 1a).
+SN, covering approximately 18,450 km², and ST, spanning around 20,452 km², represent diverse ecological and urban landscapes, ideal for assessing large-scale segmentation model performance (Figure 3 in Appendix).
+The 4-channel (RGBi) DOP20 imagery, with a spatial resolution of 20 cm per pixel, enables precise tree crown delineation.
+The DeepTrees model identified approximately 218.7 million individual tree crowns—137.3 million in SN and 81.4 million in ST (Figure 1a), and derived allometrical traits of crown area and crown spread (Figure 1b).
+A clear positive correlation between crown spread and crown area is observed, although deviations from the expected linear trend highlight segmentation uncertainties, particularly for small and irregularly shaped crowns.
+This variation underscores the need for improved methods to capture crown geometry and supports the use of allometric checks for quality assessment.
 
@@
 \includegraphics[width=0.8\textwidth]{figures/figure1.png}
-\caption{a.
-Spatial distribution of individual tree crown segmentation polygons derived from multispectral DOP20 imagery across the German federal states of Sachsen (SN, green) and Sachsen-Anhalt (ST, blue), totaling 218.7 million predicted crowns.
-b.
-Relationship between normalized crown area (x-axis) and estimated crown spread (y-axis) for all segmented trees.
-A clear positive correlation is observed, although deviations from the expected linear trend highlight segmentation uncertainties, particularly for small and irregularly shaped crowns.
-This variation underscores the need for improved methods to capture crown geometry and supports the use of allometric checks for quality assessment.}
+\caption{(a) Spatial distribution of individual tree crown segmentation polygons derived from multispectral DOP20 imagery across the German federal states of Sachsen (SN, green) and Sachsen-Anhalt (ST, blue), totaling 218.7 million predicted crowns.
+(b) Relationship between normalized crown area (x-axis) and estimated crown spread (y-axis) for a subset of the segmented trees from Halle (Saale), ST.}
 \end{figure}
+```
+:::
 
@@
 Among these challenges is phenology, a moving target: the same forest can look drastically different between leaf-on summer imagery and leaf-off winter scenes.
-Models trained on a single phenological period often struggle in another, yielding inconsistent segmentation accuracy across the year [@sapkota2024; @liu2024].
-For example, a canopy delineation that performs well on lush summer foliage may under-segment sparse autumn crowns or miss bare branches in winter.
+Models trained on a single phenological period often struggle in another, yielding inconsistent segmentation accuracy across the year [@sapkota2024; @liu2024; @garnot2025deep].
+For example, a canopy delineation that performs well on lush summer foliage may under-segment sparse autumn crowns or miss bare branches in winter (Figure 3a).
 Such season-driven variability not only degrades model performance but also complicates validation – a one-shot model might appear accurate in one season and fail in the next, raising questions about how and when accuracy should be assessed.
@@
 \end{figure}
+```
+:::
 
@@
 
-Validating large-scale tree segmentation models is challenging due to seasonal variability, diverse terrain and illumination conditions, inconsistent ground truth data, ambiguous evaluation criteria, and scale-dependent performance.
+Validating large-scale tree segmentation models is challenging due to seasonal variability, diverse terrain (Figure 3b,c)[@copernicus_copdem30, @eea_corine2020] and illumination conditions, inconsistent ground truth data, ambiguous evaluation criteria, and scale-dependent performance.
 These factors significantly affect the reliability and practical value of AI-generated tree maps for forestry, ecology, and urban planning.
@@
 \begin{credits}
-\subsubsection{\ackname} DeepTrees is part of the [“DeepTrees: Deep-Learning based spatiotemporal tree inventorying and monitoring from public orthoimages”](https://deeptrees.de) project, funded by the Integration Platform "Sustainable Future Land Use" at Helmholtz-Centre for Environmental Research – UFZ within the Programme oriented Funding (PoF) period IV of the Helmholtz Program "Changing Earth – Sustaining our Future", Topic 5 "Landscapes of the Future.
+\subsubsection{\ackname} DeepTrees is part of the \href{https://deeptrees.de}{DeepTrees: Deep-Learning based spatiotemporal tree inventorying and monitoring from public orthoimages} project, funded by the Integration Platform "Sustainable Future Land Use" at Helmholtz-Centre for Environmental Research – UFZ within the Programme oriented Funding (PoF) period IV of the Helmholtz Program "Changing Earth – Sustaining our Future", Topic 5 "Landscapes of the Future.
 This repository is based on the work described in Freudenberg et al.
